---
title: Competitive interactions are disrupted by habitat alteration in the Neotropics
output:
  bookdown::word_document2: default
bibliography: References.bib
csl: science.csl

---

__Authors:__ Anikó B. Tóth^1^*, Andrew P. Allen^2^, S. Kathleen Lyons^3^, John Alroy^2^

__Affiliations:__

^1^ Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW, Sydney, NSW 2052, Australia.

^2^ Department of Biological Sciences, Macquarie University, New South Wales 2109, Australia.

^3^ School of Biological Sciences, University of Nebraska-Lincoln, Lincoln, Nebraska 68588, USA.

*Correspondence to aniko.toth@unsw.edu.au

__Abstract:__ Wholesale habitat destruction disrupts biotic interactions. Although this process may have far-reaching consequences, assemblage-wide studies of this phenomenon on landscape scales are uncommon and beset by logistical hurdles. Here, we use a novel approach to assess interactions in Neotropical bird and bat assemblages. We ask whether spatial patterns of co-occurrence among species pairs that compete for food resources differ between altered and intact habitats. We find that competing species pairs of both taxa are more spatially segregated on average in altered habitats than intact habitats. The shift is not driven by diversity losses but by changes in spatial relationships of recurring species pairs. Our results suggest that altered habitats fail to support the coexistence of diverse competitive interactions, reversing patterns observed in the wild.

__One sentence summary:__ Coexistence of competing species of Neotropical bats and birds is reduced in human-altered habitats.

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(igraph)
library(reshape2)
library(vegan)
library(sp)
library(stringi)
library(parallel)
library(magick)

knitr::opts_chunk$set(include = FALSE)

inline_hook <- function(x) {
  if (is.numeric(x)) {
    format(x, digits = 2)
  } else x
}

knitr::knit_hooks$set(inline = inline_hook)

## Load helper functions
source('../Code/HelperFunctions.R')
## Prep raw data
source('../Code/Data_Prep.R')

```


__Main text:__
Biological communities depend on a balance of interspecific interactions. Human disturbance may rewire interactions in unpredictable ways [@Bartley2019], and this can have detrimental consequences for ecosystems and people. However, empirical studies quantifying how biotic interactions change under degraded conditions are uncommon. Evidence that human disturbance can decouple predator-prey interactions [@Rodewald2011], increase competition for pollinators [@Holzschuh2011], and simplify or break apart interaction networks [@Haddad2000; @DeAssisBomfim2018] has been gathered mainly at local scales on systems with only a few species due to logistical hurdles. Broad patterns documenting changing interactions in entire assemblages have rarely been revealed, except for systems where interaction data has been laboriously compiled over decades, such as plant-pollinator networks [@Dore2021]. These are nonetheless beset by sampling inconsistencies [@Dore2021].

Biotic interactions are one of the four main drivers of species distributions [@McGill2010]. Thus, it is crucial to understand how assemblage-level interaction dynamics change in various ecological contexts [@Song2020], to facilitate conservation of species and ecosystems in an uncertain future. For instance, species pairs can exhibit stable or unstable population dynamics depending on the presence or absence of an apex predator [@Karakoc2020], or switch from facilitation to competition along a climate gradient [@Bimler2018]. Competitive interactions can result in coexistence or exclusion, depending on fitness, niche disparity, and priority effects of the species involved [@Grainger2019]. Human habitat alteration has strong potential to influence these processes by extirpating apex predators, changing the climate, or concentrating resources, thereby shifting the dynamics of biotic interactions, often without necessarily adding or removing species [@Tylianakis2017].

Although this interpretation is still contentious [@McGill2010; @Araujo2013; @Ezard2016], it is increasingly recognized that biotic interactions can scale up to influence community assembly at much larger spatiotemporal scales than that of individual encounters [@Araujo2007; @Pollock2014; @Fraser2020]. For example, competition and facilitation can influence the modern distribution of taxa at landscape scales [@Gotelli2010], at continental scales over millennia [@Toth2019], and drive the diversification and extinction of entire clades over millions of years [@Sepkoski1984; @Alroy2010; @Fraser2020]. Human disruption of interactions can therefore have far-reaching consequences.

We demonstrate how human habitat alteration influences one type of interspecific interaction (food competition) through its effects on spatial associations of species pairs (meaning that species aggregate or co-occur more often than expected, vs. segregating, co-occurring less often than expected). We introduce a statistical model that elucidates how associations between individual pairs of species collectively result in assemblage-level spatial patterns. We apply the model to assemblages of Neotropical bats and birds downloaded from the Ecological Register [@Alroy2015] consisting of `r sum(PAn[["bat"]])` occurrences of bats and `r sum(PAn[["bird"]])` occurrences of birds at `r ncol(PAn[["bat"]])` and `r ncol(PAn[["bird"]])` sites, respectively. The sites in the Register are classified into two levels of degradation (altered vs. intact habitats; Fig. S1). Our analysis demonstrates that habitat degradation influences spatial patterns of co-occurrence among competitors at regional to continental scales by altering – or even reversing – the average outcome of food competition across regional assemblages. 

```{r biogeocorr, echo = FALSE}
#### Match Biogeography ####
# separate altered and intact sites
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites])) # separate intact
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites])) # and altered sites

coords <- sitedat %>% select(taxon, p.sample, latitude, longitude) %>% distinct() %>% split(.$taxon)

# Remove intact or altered sites that are not near a site of the other type. This is done to ensure the two sets have similar biogoegraphical distributions.
coords1 <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2 <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])
keep <- map2(coords1, coords2, matchbiogeo) %>% map(unlist)

PAn <- map2(PAn, keep, function(x, y) return(x[,as.character(y)]))
PAn <- map(PAn, clean.empty, minrow = 1) # remove any species that now have no occurrences

# recalculate alt/unalt split from new PAn
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites]))
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites]))

# coords of sites we kept, for plotting later.
coords1.keep <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2.keep <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])

```


```{r richness, include = FALSE}
###### Richness - run using raw abundance data ####
# can only be run on raw data because it requires a singleton count of abundances.
rich <- list(cJ1 = map(PAn, map_dbl, cJ1rich) %>% map(~split(., names(.) %in% unalt_sites)) %>% # corrected 1st order jackknife
               map(map, cbind) %>% map(map, data.frame), 
             chao = map(PAn, map_dbl, chao1) %>% map(~split(., names(.) %in% unalt_sites)) %>%  # chao1
               map(map, cbind) %>% map(map, data.frame) , 
             fa = map(PAn, map_dbl, fisher.alpha) %>% map(~split(., names(.) %in% unalt_sites)) %>%  # fisher's alpha
               map(map, cbind) %>% map(map, data.frame)) %>% 
  map(map, bind_rows, .id = "status") %>% map(bind_rows, .id = "taxon") %>% bind_rows(.id = "metric") %>% 
  setNames(c("metric", "taxon", "status", "richness")) %>% 
  mutate(status = recode(status, `FALSE` = "Altered", `TRUE` = "Intact"))

# Summary
rsumm <- rich %>% group_by(metric, taxon, status) %>% summarise(mean.rich = mean(richness, na.rm = T), median.rich = median(richness, na.rm = T))
# Significance
rsig <- rich %>% group_by(metric, taxon) %>% summarise(p=wilcox.test(richness~status, paired=FALSE)$p.value, 
                                               W=wilcox.test(richness~status, paired=FALSE)$statistic)

```


```{r betacomp, echo=FALSE, message=FALSE, warning=FALSE}

#### Beta diversity and composition analyses #####
PAnb<- tobinary(PAn)

PA <- PAnb  ## can run analyses with binary or abudance data

# with bray-curtis index
dist <- map(PA, ~t(.)) %>% map(vegdist, method = "jaccard")
# with Ochiai index or cosine similarity
dist <- map(PA, as.matrix) %>% map(cosine) %>% map(as.dist, upper = F) %>% map(~return(1-.))

# beta diversity using betadisper
beta <- map2(dist, PAn, function(x, y) betadisper(x, group = colnames(y) %in% unalt_sites) %>% anova)

# compositional change using adonis
data <- map(PA, ~t(.x) %>% data.frame() %>% rownames_to_column("p.sample") %>% mutate(ID = colnames(.x) %in% unalt_sites))
comp1 <- map2(dist, data, function(x, y) adonis(x~ID, data = y, permutations = 10000))

# compositional change using canonical correspondence analysis
dat <- map(data, select, contains("_"))
comp2 <- map2(dat, data, function(x, y) cca(x~ID, data = y) %>% anova.cca)

```

A suite of classical community analyses using several metrics produces inconsistent results on whether community patterns differ between altered and intact sites [@Supplement]. Specifically, birds have significant changes in community composition without changes in richness or beta diversity (Table S1). Bats have no significant disparity in beta diversity but possible changes in richness and composition, depending on the analysis and metric used (Table S1). We ran a multidimensional scaling analysis which demonstrates visually that compositional dissimilarities between altered and intact habitats are not striking (Fig. 1). In sum, classical analyses detect possible differences between habitat types, but they are not necessarily robust or informative about shifts in processes (as opposed to patterns) such as competitive exclusion and resource partitioning.

```{r npairs, message=FALSE, warning=FALSE, include=FALSE}
tables <- PAnb %>% map(~t(.)) %>% map(as.data.frame) %>% map(~split(., f = rownames(.) %in% unalt_sites)) %>% 
  map(map, ~t(.)) %>% map(map, clean.empty) %>% purrr::map(setNames, c("altered", "unaltered"))

original <- map(tables, map, ~expand.grid(Sp1 = rownames(.x), Sp2 = rownames(.x))) %>% map(map, diet_cat, spp, related = TRUE) %>% map(bind_rows, .id = "status") %>% bind_rows(.id = "taxon") %>% na.omit() %>% filter(Sp1 != Sp2) %>% select(taxon, Sp1, Sp2, diet.match) %>% distinct() %>% group_by(taxon) %>% summarise(pairs = n(), related = length(which(diet.match == "Related")), pct_related = 100*related/n())

```

To characterize the effect of food competition on community assembly, we instead considered every possible pairwise combination of species within each taxon (`r original %>% filter(taxon =="bat") %>% pull(pairs)` bat and `r original %>% filter(taxon =="bird") %>% pull(pairs)` bird pairs). Each pair was classified as potentially competing (henceforth: competing) or not competing (henceforth: control) based on basic dietary guild data (Fig. 2) from the Ecological Register [@Supplement]. Competing pairs shared dietary guilds while control pairs shared no dietary sources. For simplicity, we excluded pairs with partially overlapping diets (`r original %>% filter(taxon =="bat") %>% pull(pct_related)`% of bat and `r original %>% filter(taxon =="bird") %>% pull(pct_related)`% of bird pairs). The intersection of diet and habitat type produced four experimental groups: competing pairs in intact sites, control pairs in intact sites, competing pairs in altered sites, and control pairs in altered sites. We fit a hierarchical Bayesian model [@Supplement] that yields group-level estimates ($g$) for the average ($\bar{\theta_{g}}$) and standard deviation ($\sigma_{g}$) of a spatial aggregation parameter ($\theta$) that indexes pairwise co-occurrence using Fisher’s noncentral hypergeometric distribution [@Fog2008]. A negative value of $\theta$ indicates spatial segregation of the species pair, whereas a value of approximately 0 indicates independent patterns of occupancy for the pair and a positive value indicates spatial aggregation. Models were run on all pairs (“full”), then repeated on those exceeding the median incidence rate in their taxon and habitat type (“common-only”) to reduce uncertainty due to low occurrences. Posterior distributions of competing pairs were compared to the control group in each habitat for each taxon using 95% high density intervals (HDIs) on 1,200 posterior samples (Figs 3-4, Table S2). Though sharing diet assignments clearly does not guarantee that a species pair competes for food, this design ensures that—barring errors in assignment—all pairs that do compete for food are analyzed in the same group. The inclusion of potentially many pairs that do not truly compete in this group reduces the power of the model but should produce conservative estimates of differences between groups. The pairwise approach results in non-independence of data within each habitat, so the significance of the results cannot be formally evaluated [@Supplement]. Based on these two considerations, it is difficult to say whether the underlying effect is large in absolute terms, which is a source of uncertainty in our results.

Overall, birds tend to segregate ($\theta < 0$) while bats tend to aggregate ($\theta > 0$; Figs 3-4). This observation represents fundamental difference in the ecology of birds and bats. Bats exhibit limiting morphological similarity [@Kingston2000], share roosts [@Swift1983], display varied foraging behaviours [@Swift1983; @Denzinger2013], and micro-partition resources, habitats, and foraging times [@Aldridge1987]. Thus, they can co-exist locally. By contrast, Neotropical bird species are well known for interspecific aggression [@Freeman2016a], separating out along elevational gradients [@Freeman2016a], and competitive exclusion [@Remsen1995]. Even control bird pairs are far less likely to aggregate strongly than control bat pairs (compare x-axis in Figs. 3-4 $\theta$ panels), suggesting that natural beta diversity and abiotic niche specialization are much more pronounced in birds than bats. Nonetheless, competition for food affects the spatial community structure of both taxa regardless of habitat as demonstrated by offsets in the posterior distributions for $\bar{\theta{g}}$ and/or $\sigma_{g}$ of competing and control groups (Figs. 3-4, Table S2). Competing pairs consistently exhibit higher variance than control pairs, supporting the conclusion that spatial patterns are more variable among species that share food sources (Fig. S2).

In intact habitats, we find no compelling evidence for assemblage-wide incidence of diet-based competitive exclusion in either taxon. Competing pairs coexist equally (full models) to or more readily (common-only models) than control pairs on average (i.e., $\overline{\theta{comp}} >= \overline{\theta{contr}}$) for both bats (Fig. 3A-B) and birds (Fig. 4A-B). Because bats tend to aggregate and birds tend to segregate, this translates into stronger attraction of competing bats than control bats and to weaker repulsion of competing birds than control birds. Competing variances are also higher than in the control group, suggesting that mutual food sources encourage resource sharing and partitioning in the wild. In both taxa, this pattern changed in altered habitats.

For bats, the offset in $\bar{\theta{g}}$ between competing and control pairs disappears in altered habitats, and variances are higher for both competing and control groups. In the common-only model (Fig. 3B), there is some evidence that competing pairs are less likely to coexist than control pairs, though 95% HDIs are overlapping. While aggregation may still occur in some pairs, exclusion is much more common for competing pairs under altered than intact conditions. Full models (Fig. 3A) and common-only models (Fig. 3B) agree in this respect, but full models also exhibit lower $\bar{\theta{g}}$ for both altered groups. For bats, natural coexistence is therefore less common in altered habitats, particularly for rarer species, and irrespective of competitive status.

For birds, the resource partitioning observed in intact habitats is reversed, with control pairs exhibiting dramatically higher posteriors ($\bar{\theta{g}} > 0$, uncharacteristically for birds) and lower variance, while competing pairs remain just above intact levels of $\bar{\theta{g}}$ (common-only models Fig. 4B) or well below them (full models Fig. 4A) in altered habitats. This suggests that altered habitats supply more opportunities for co-occurrence than intact habitats (for instance, through resource concentration or niche construction), but competing pairs are unable to capitalize on these opportunities. This causes them to instead exhibit exclusion compared to control pairs. The altered control group in the full model returned an unconverged posterior distribution, but the underlying $\bar{\theta{g}}$ of this group is consistent with our interpretations no matter where it falls in the wide possible range. 

To discover whether compositional differences between habitats were responsible for our results, we re-run our models (both “full” and “common-only”) while excluding species that were not sampled in both altered and intact habitats ("no-turnover" models) [@Supplement]. In bats there is no qualitiative change (Fig. 3C-D), meaning that recurring pairs change their spatial behavior between habitats. In birds, the patterns within habitat types are conserved but the absolute differences between habitat types are reduced (Fig. 4C-D), suggesting that compositional turnover and changes in the spatial associations of recurring pairs both play a role [@Supplement]. 

We find differences in the community structures of bats and birds in intact habitats, but for both taxa there was higher coexistence of competing pairs than expected, suggesting that resource partitioning occurs most frequently instead of competitive exclusion. Patterns changed in altered habitats. Altered habitats afford fewer opportunities for the coexistence of bats but more opportunities for the coexistence of birds, especially those sampled exclusively in altered habitats. Competing pairs, however, exhibited equal or lower coexistence than control pairs under altered conditions, suggesting that competitive exclusion is more common. Thus, altered habitats are unable to consistently support varied local communities of competing species through partitioning, as seen in intact habitats. The no-turnover analyses indicate that these effects are largely attributable to changes in spatial behavior, and not merely to differences in species composition.

Overall, habitat alteration reduces or reverses the effect of competition on community structure in both bats and birds. Our results add to mounting evidence that human activity changes community structure and interactions without necessarily adding or removing species, and further corroborate that this can play out over landscape scales. Though diversity loss is known to influence ecosystem services [@Winfree2018], analogous studies examining the relationships between biotic interactions and functionality of ecosystems are rare (but see e.g. [@Cardinale2002]). Consequences of shifting competition outcomes are therefore poorly understood, but negative scenarios are easy to envision. For example, population control of a destructive pest by a competing species can not take place if the two species are unable to coexist. Our results shed light on the elusive relationship of biotic interactions on community assembly processes, and our approach helps pave the way for future research seeking to understand how interaction outcomes change across various contexts and at large spatial scales.


__Acknowledgments:__ We thank the Macquarie University paleobiology lab for discussions that improved this paper. 

__Funding:__ ABT was partially supported by an IMQRES scholarship at Macquarie University.

__Author contributions:__ ABT and JA designed the study, JA curated the data and provided statistical expertise, ABT analyzed data, produced the figures, and wrote the paper. ABT and APA designed, coded, and tested the model. SKL helped design supporting analyses. All authors edited the manuscript.

__Competing Interests:__ The authors declare no competing interests.

__Data and materials availability:__ All R workflows and cleaned datasets used for this analysis are available at https://github.com/anikobtoth/HabitatAlteration.

### Figures

```{r echo=FALSE, fig.height=4, fig.width=9, message=FALSE, warning=FALSE, include=T}

# Ordination
dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
#dist <- map(PA, ~t(.)) %>% map(vegdist, method = "bray")
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "p.sample"))

bat <- image_read("bat_pic.png") %>% image_fill('none') %>% as.raster()
bird <- image_read("bird_pic.png") %>% image_fill('none') %>% as.raster()

#PCoA plot
  p1 <- ggplot(ord[[1]], aes(x = X1, y = X2, col = status)) + geom_point(size = 2.6) + labs(x = "PCoA 1", y = "PCoA 2", col = "alteration") + theme(axis.text = element_text(size = 12), legend.text = element_text(size = 12), legend.title =  element_text(size = 12, face = "bold")) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77")) +
    annotation_raster(bat, -0.45, -0.2, 0.27, 0.42)#+ geom_point(aes(x = mean(X1), y = mean(X2), col = status), pch = 3, size = 3)
  
  p2 <- ggplot(ord[[2]], aes(x = X1, y = X2, col = status)) + geom_point(size = 2.6) + labs(x = "PCoA 1", y = "PCoA 2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77"), labels = c("Altered", "Intact")) +
     annotation_raster(bird, -0.45, -0.25, 0.25, 0.44) #+ geom_point(aes(x = mean(X1), y = mean(X2), col = status), pch = 3, size = 3)
  
  p <- plot_grid(p1+theme(legend.position = "none"), 
                 p2+theme(legend.position = "none", 
                axis.title.y = element_blank()), ncol = 2, labels = c("A","B"), align = "lr")
  plot_grid(p, get_legend(p1), rel_widths = c(1, .2))

```

__Figure 1.__ Multidimensional scaling analysis of altered and intact sites for bats (A) and birds (B) based on the Ochiai dissimilarity index [@Hubalek1982]. 


__Figure 2.__ Conceptual figure demonstrating the experimental design. (A) All pairwise combinations of species are classified as potentially competing or not competing for food based on basic dietary information. (B) A distribution of $\theta$ is calculated for each pair, shown as an adjacency matrix. From these distributions, (C) the cumulative posterior distribution of group means $\bar{\theta{g}}$ is derived and compared. 

```{r echo=FALSE, include = T, warnings = FALSE}
colors <- c("#FFC2AD", "#DF8768", "#83C5BE", "#006D77")
parameter.labels <- c(omega = "group θ", sigma = "variance (σ)")
ann <- data.frame(value = -Inf, lab = LETTERS[1:4], 
                  parameter = "omega", 
                  model = factor(c("full", "common", "full noTurnover", "common noTurnover"), levels = c("full", "common", "full noTurnover", "common noTurnover")), 
                  status_dietmatch = "Unaltered_Different")

specs <- list(geom_density(lwd = 1, alpha = .4),
              facet_grid(model~parameter, scales = "free_x", labeller = labeller(parameter = parameter.labels)), 
              scale_color_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")), 
              scale_fill_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")), 
              labs(fill = "", col = "", x = "Value", y = "Density"), 
              theme(text = element_text(size = 12)),
              geom_text(data = ann, y = Inf, label = ann$lab, col = "black", fontface = 2, hjust = -1, vjust = 1))
```

```{r echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE, include=T}
load("../Results/Omega/interaction/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
bat_common <- out
load("../Results/Omega/interaction/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
bat_full <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
bat_common_noTurnover <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
bat_full_noTurnover <- out

out <- bind_rows(list(full = bat_full, 
                      common = bat_common, 
                      full_noTurnover = bat_full_noTurnover, 
                      common_noTurnover = bat_common_noTurnover), .id = "model") %>% 
  filter(!parameter == "deviance") %>%
  mutate(model = recode(model, full_noTurnover = "full noTurnover", common_noTurnover = "common noTurnover"))

out$model <- factor(out$model, levels = c("full", "common", "full noTurnover", "common noTurnover"))


ggplot(out, aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + specs


```

__Figure 3.__ $\bar{\theta{g}}$ and $\sigma_{g}$ posteriors for bat models, specifically (A) all bats, (B) all bats having occurrences that exceed the median occurrence count in the respective habitat, (C) and all bats that occurred in both intact and altered habitats, and (D) all bats having occurrences that exceed the median occurrence in both intact and altered habitats. C and D represent the models in which altered and intact faunas are composed of the same suite of species (no-turnover models).

```{r echo=FALSE, fig.height=7, fig.width=7, include = T, warning = FALSE, message = FALSE}

load("../Results/Omega/interaction/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
bird_common <- out
load("../Results/Omega/interaction/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
bird_full <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
bird_common_noTurnover <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
bird_full_noTurnover <- out


out <- bind_rows(list(full = bird_full, 
                      common = bird_common, 
                      full_noTurnover = bird_full_noTurnover, 
                      common_noTurnover = bird_common_noTurnover), .id = "model") %>% 
  filter(!parameter == "deviance") %>%
  mutate(model = recode(model, full_noTurnover = "full noTurnover", common_noTurnover = "common noTurnover"))

out$model <- factor(out$model, levels = c("full", "common", "full noTurnover", "common noTurnover"))

ggplot(out, aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + specs


```

__Figure 4.__ $\bar{\theta{g}}$ and $\sigma_{g}$ posteriors distributions for bird models. Panels as in Fig. 2.


