---
title: The effect of food competition on community assembly in Neotropical bats and birds is reversed by habitat alteration
author:
  - "Anikó B. Tóth"
  - "Andrew P. Allen"
  - "S. Kathleen Lyons"
  - "John Alroy"
output:
  bookdown::word_document2: default
bibliography: References.bib
csl: proceedings-b.csl

---

^1^ Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW, Sydney, NSW 2052, Australia.
^2^ Department of Biological Sciences, Macquarie University, New South Wales 2109, Australia.
^3^ School of Biological Sciences, University of Nebraska-Lincoln, Lincoln, Nebraska 68588, USA.


__Abstract__

Anthropogenic habitat destruction is one of the greatest threats facing animal communities. To advance conservation and management initiatives, it is crucial to understand how habitat alteration changes the way communities assemble. Increasing evidence points to the role of biotic interactions in structuring communities even on large spatial scales, and human activities may disrupt these processes indirectly. However, we do not yet have widely applicable, robust methods for quantifying the effects of biotic interactions in biodiverse assemblages. Here, we present a bayesian approach that detects and quantifies the role of diffuse competition in shaping large assemblages, specifically, whether or not it tends to result in exclusion. Using our approach, we examine the outcome of food competition in neotropical birds and bats, and compare the results for assemblages collected from human-altered and unaltered habitats. Food competition significantly influenced community assembly of bats and birds, and its effects were stronger in altered than unaltered habitats even though richness and composition were not significantly different. Most competing species pairs exhibited stronger segregation in unaltered habitats and stronger aggregation in altered habitats in contrast to species that do not compete. By contrast, the minority of competing species that prefer unaltered habitats exhibited the opposite pattern in both bats and birds, segregating strongly in altered habitats even while little or no effect of food competition was observed in unaltered conditions. Our results suggest that alteration reduces competitive exclusion, possibly by providing abundant food sources at predictable places and times, except for species intolerant of human disturbance, which are rarely able to persist in altered conditions when a competing species is present. Habitat alteration therefore strongly affects community structure via changes in competitive outcomes, nullifying and often reversing patterns observed in the wild.


```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(igraph)
library(reshape2)
library(vegan)
library(sp)
library(stringi)
library(parallel)

knitr::opts_chunk$set(include = FALSE)

## Load helper functions
source('../Code/HelperFunctions.R')
## Prep raw data
source('../Code/Data_Prep.R')

# species metadata
#load("../Data/Species_metadata.RData")
#load("../Data/cosmo_species_cosmopolitan_score.RData") # cosmopolitan category 
# Site data #
#load("../Data/sitedat_metadata.Rdata")

## Richness results
load("../Results/cj1_chao_richness_raw.RData")

taxa = c("bat", "bird")

unalt_sites <- sitedat[sitedat$altered_habitat == "",]$siteid

```


### Main text: 

#### Introduction

How does human habitat alteration disrupt biological assemblages? With rapidly changing global conditions, this question is on the forefront of many urgent research agendas, spanning conservation, restoration, environmental science, ecology, and paleoecology. It is increasingly recognised that the effects of human disturbance extend many thousands of years into the past [@Lyons2016; @Maezumi2018], across the entire planet [@Sanderson2002; @Ellis2008], and may continue to be felt tens of millions of years into the future [@Novacek2001; @Alroy2008]. 

Human impacts on the natual drivers of community assembly, such as climate change [@IPCC2007], habitat alteration [@Alroy2017], engineering of dispersal barriers [@Cui2018], and changes in interaction networks [@deAssisBomfim2018], have been a focus of intense research interest. Aside from the straightforward introduction or removal of species, the last of these is probably the most poorly understood, with many researchers still working to understand the dynamics of interaction systems without even considering human impacts explicitly [@Baiser2019]. Nonetheless, there is evidence that human disturbance can decouple predator-prey interactions [@Rodewald2011], increase competition for pollinators [@Holzschuh2011], and simplify or break apart interaction networks [@Haddad2000; @deAssisBomfim2018].

Though the subject has long been contentious among both ecologists [@McGill2010; @Araujo2014] and palaeontologists [@Ezard2016; @Marshall2016], evidence has long accumulated that the effects of biotic interactions scale up to measurably influence community assembly at much larger spatial scales [@Araujo2007; @Pollock2014; @Fraser2021 in press] than the scale of individual encounters would suggest. The same appears to be true for biotic processes occurring over geological timescales, as mounting evidence demonstrates that interactions such as competition and facilitation can be instrumental in structuring spatial patterns over thousands of years [@Toth2019], as well as the diversification and extinction of entire clades on timescales of millions of years [@Sepkoski1984; @Alroy2010]. Though many interaction networks have been documented in detail, these are usually limited to single locations or small extents, so the way in which this upscaling unfolds is still not well understood.

One obvious reason for the limited understanding of biotic interactions is that they are costly and difficult to monitor directly. Quantitative methods that have the capability to distinguish the effects of biotic interactions from other drivers are therefore becoming crucially important, and indeed have already begun to yield promising results [@Fraser2021 in press]. Such methods rely on large datasets of species occurrence and ecological traits that are becoming widely available and were therefore not truly feasible until very recently. The question of shifting biotic interactions is as critical as it is elusive. The degree to which biotic interactions have been disrupted is one of the five criteria proposed for assessing the risk level of ecosystems and has been adopted by the IUCN Red List of Ecosystems [@Keith2015]. Humans produce pollution, change the spatial allocation of resources, genetically modify organisms, and engineer novel ecological niches, demonstrating a vast potential to affect the dynamics of biotic interactions and thus the composition of assemblages living in proximity to human-altered areas, all without actually adding or removing species.

Herein, we present a Bayesian method that quantifies the effect of diffuse biotic interactions on large assemblages. Specifically, it identifies whether interacting species in large assemblages (e.g. those competing for food) tend to exhibit coexistence, exclusion, both, or neither.  The model allows users to examine individual pairs or groups of pairs for their contributions to the overall pattern, thus moving toward a better understanding of how biotic interactions scale from individual interspecific interactions to landscape-scale spatial patterns. We demonstrate the model by applying it to assemblages of neotropical bats and birds and compare the outcome of diffuse food competition between altered and unaltered habitats. Our analysis sheds light on how human habitat disturbance influences spatial patterns by altering the course of competitive interactions across entire assemblages.


#### Materials and Methods

```{r, echo = FALSE}
unalt_sites <- sitedat %>% filter(status == "Unaltered") %>% pull(sample.no) 
```

Abundance data were dowloaded from the Ecological Register (http://ecoregister.org) on 7th June 2020 for mistnetted bats and birds from the Neotropics, defined as all areas in the western hemisphere south of the Tropic of Cancer (23.44˚N). We also dowloaded Register metadata for the species and sites in the abundance tables, including coordinates and habitat alteration category for sites and dietary guild assignments for species. Guild assignments were based on a large set of recently published individual papers. Sampling and reporting was standardised across the samples, as all were mist-netted samples with consistent mesh size, harp net samples were excluded, and almost all nets were placed at ground level. Every sample consisted of a minimum of 20 individuals and the vast majority of sites had more than 50, with the average number of individuals per sample being `r mean(colSums(PAn[["bat"]]))` for bats and `r mean(colSums(PAn[["bird"]]))` for birds.  Samples were combined if they were within the same 11.1 km equal-area grid cell, sourced from the same study, and represented the same habitat and alteration category. The combined data consisted of a total of `r sum(PAn[["bat"]])` bat and `r sum(PAn[['bird']])` bird individuals from `r ncol(PAn[['bat']])` and `r ncol(PAn[['bird']])` sites, respectively. Maps of the sites are plotted in Figure S1.


```{r, include = FALSE}
#### Change to presence-absence ####
PAn<- tobinary(PAn)
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites])) # separate unaltered
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites])) # and altered sites

#### Match Biogeography ####
meta <- sitedat %>% split(.$taxon)
coords <- map(meta, select, c(sample.no, latitude, longitude)) 

# Remove unaltered or altered sites that are not near a site of the other type. This is done to ensure the two sets have similar biogoegraphical distributions.
coords1 <- purrr::map2(coords, PAnu, function(x, y) x[x$sample.no %in% colnames(y),])
coords2 <- purrr::map2(coords, PAna, function(x, y) x[x$sample.no %in% colnames(y),])
keep <- map2(coords1, coords2, matchbiogeo) %>% map(unlist)

PAn <- map2(PAn, keep, function(x, y) return(x[,as.character(y)]))
PAn <- map(PAn, clean.empty, minrow = 1) # remove any species that now have no occurrences

# recalculate alt/unalt split from new PAn
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites]))
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites]))

# coords of sites we kept, for plotting later.
coords1.keep <- purrr::map2(coords, PAnu, function(x, y) x[x$sample.no %in% colnames(y),])
coords2.keep <- purrr::map2(coords, PAna, function(x, y) x[x$sample.no %in% colnames(y),])

```

*__Richness and Beta diversity.__* We estimated richness for bats and birds in altered and unaltered sites using a corrected first-order jackknife (cJ1) [@Alroy2020] and Chao1 [@Chao1984] on the raw abundance data for all sites. The status of each site was based on the 'altered.habitat' field in the site metadata, which is empty if the site is unaltered and filled with a land use category if the site is altered. Altered sites included cropland, disturbed forest, fragment, pasture, plantation, secondary forest, inhabited area, and combined categories [@Alroy2017], and none of them had enough sites to be evaluated individually. The abundance data were converted to a presence-absence format. 
The spatial coverage of altered sites was a subset of the larger spatial area covered by unaltered sites. To minimize the effects of biogeography on our results, we removed unaltered sites that were farther from the nearest altered site than the most isolated altered site was from its nearest unaltered site (see supplement), resulting in comparable spatial coverage for both site types. The resulting dataset had `r sum(PAn[["bat"]])` occurrences of bats and `r sum(PAn[["bird"]])` occurrences of birds at `r ncol(PAn[["bat"]])` and `r ncol(PAn[["bird"]])` sites, respectively. We then calculated the beta diversity between pairs of sites for bats and birds separately, separating them into altered and unaltered groups. We used the Ochiai similarity index as the similarity measure for beta diversity, and transformed the results to z-scores to achieve normality for significance testing. Finally, we ran a principal coordinates analysis (PCoA) on the sites using the R function *cmdscale* in the stats package to check for compositional shifts between altered and unaltered habitats. We tested for significant shifts in the site centroids on the two primary PCoA axes using a simple permutation test. 

*__Biotic Interactions.__* We used dietary information for bats and birds from the Ecological Register to analyse the effects of food competition on patterns of co-occurrence (Table S1). We selected all pairs with complete dietary information and classified them as competing or non-competing. Competing pairs shared primary and secondary sources of food, but we allowed the order of these to be reversed (e.g., a frugivore-nectarivore and a nectarivore-frugivore would be equated). Non-competing pairs had no mutual food sources. Pairs with mutual and nonmutual food sources (e.g., a frugivore and an insectivore-frugivore) may experience a complex mix of effects and were therefore removed from the analysis. Although species in the same dietary guilds are not necessarily competing, pairs with high dietary overlap have a nonzero probability of experiencing competition for food, while pairs in differing guilds have no chance of competing for food. Therefore, the co-occurrence patterns of competing and non-competing groups should be different if food competition plays a role in structuring the assemblage. 

Note that any type of diffuse interaction can theoretically be represented with this approach, as long as some sort of a priori evidence exists that there may be an interaction. For instance, a more powerful version of the competition model could be created by placing only pairs with the same diet and similar body size in the competing group. Conversely, a plant-pollinator network could be decomposed into groups of pairs that facilitate one another (plant-pollinator pairs), do not interact, and even compete (plant-plant pairs that compete for pollinators or pollinator-pollinator pairs that compete for pollen). In each case, the effect of the competing group(s) is evaluated against the group of pairs that does not exhibit the interaction(s) in question. Unlike most analyses in the recent literature, pairs that lack an arbitrarily significant interaction are not removed from the analyses, but instead used to help estimate the true differences between groups that interact and do not interact. 

This framework also allows invesigators to quantify the effect of various external factors on the spatial outcomes of diffuse biotic interactions. In the example below, we compare assemblages from altered and unaltered habitats to evaluate the effect of habitat alteration on the outcome of food competition. However, the samples could be split up along other lines instead, such as environmental gradients, temporally before and after disturbances, etc.

Recent studies repeatedly point out that nonrandom co-occurrence should not be construed as evidence for interaction between species pairs [@Freilich2018;@Thurman2019; @Blanchet2020]. In our framework, we have independent evidence of potential interactions (diet) and examine the relationship between putative interactions and co-occurrence, using non-interacting pairs as a control. Research using such an approach is exactly what is needed to understand under what circumstances, and to what extent, interactions lead to nonrandom spatial associations, what factors affect the relationship, and how they change under duress. 

*__The Model__*
We assume that the probability that species A and B co-occur at $N_{AB}$ sites adheres to Fisher's noncentral hypergeometric distribution [@Fog2008],

\[
\tag{1a}
p(N_{AB}|\psi) = 
\frac{{N_{A^{+}} \choose N_{AB}}{N_{A^{-}}\choose N_{B^{+}} - N_{AB}} \psi^{N_{AB}} }
{\sum_{n} {N_{A^{+}}\choose n}{N_{A^{-}}\choose N_{B^{+}} - n} \psi^{n}}
\]

where $N_{A^{+}}$ and $N_{A^{-}}$ are the respective numbers of sites with species A present and absent, $N_{B^{+}}$ is the number of sites with species B present, and $\binom{n}{k} = \frac{n!}{k! (n-k)!}$ is a binomial coefficient. The denominator of this probability mass function (pmf) is the numerator summed over the range of possible values ($n$) for the numbers of co-occurring sites, $\text{max}(0,N_{A^{+}}+N_{A^{-}}-N_{B^{+}}) \leq n \leq \text{min}(N_{A^{+}},N_{B^{+}})$. This pmf has one fitted parameter,

\[
\tag{1b}
\psi = \frac{\pi_{A|B^{+}}}{\pi_{A|B^{-}}} =\frac{p_{A|B^{+}} (1 - p_{A|B^{-}})} {(1-p_{A|B^{+}}) p_{A|B^{-}}}
\]

where $\pi_{A|B^{+}}$ and $p_{A|B^{+}}$ are the respective odds and probability that species A occupies a site given that species B is present, and $\pi_{A|B^{-}}$ and $p_{A|B^{-}}$ are the respective odds and probability that species A occupies a site given that species B is absent. The value of $\psi$, and the calculated probability of obtaining $N_{AB}$ sites of co-occurrence, $p(N_{AB}|\psi)$, are independent of which member of the species pair is arbitrarily chosen as species A.


The $\psi$ parameter is defined on the interval $[0,\infty]$ because it is a ratio of odds. If $\psi < 1$, species A is less likely to occur at a site if species B is present, as would be expected if there is competitive exclusion among species or differences in species' habitat preferences. If $\psi > 1$, species A is more likely to occur at a site if species B is present, as would be expected if the species share habitat preferences and do not competitively exclude each other. Finally, if $\psi = 1$, the two species are independently distributed. When $\psi = 1$, Fisher's noncentral hypergeometric distribution reduces to the standard (i.e. centered) hypergeometric distribution, which has frequently been used to evaluate patterns of co-occurrence [@Arita2016]. To our knowledge, our study is the first to use Fisher's noncentral hypergeometric distribution for this purpose.


We adopt a hierarchical Bayesian modeling approach [@Gelman2014] to estimate averages of $\theta = \text{log}\psi$ for species pairs that have been subdivided into four groups, $g$, on the basis of habitat type (altered, unaltered) and diet (similar, distinct). At the lower hierarchical level, we model variation in $\theta_i$ for species pairs and groups as

\[
\tag{2}
\theta_i = \overline{\theta_{g(i)}} + \Delta\theta_i
\]

where $g(i)$ is a function that indexes the group assignment of each species pair $i$, and $\Delta\theta_i$ is a deviation from the group mean $\overline{\theta_{g(i)}}$. We assume that these deviations are normally distributed with means of 0,

\[
\tag{3}
p(\Delta\theta|\sigma_{g}) = \frac{1}{\sigma_{g} \sqrt{2 \pi}} e^{\Delta\theta^2 / \sigma_{g}^2}
\]

and standard deviations that vary by group, $\sigma_{g}$. At the upper hierarchical level, we assign the group averages, $\overline{\theta_g}$, normal prior distributions with means of 0 and standard deviations of 5,

\[
\tag{4}
p(\overline{\theta_g}) = \frac{1}{5 \sqrt{2 \pi}} e^{\Delta\theta^2 / 5^2}
\]

and we assign the group standard deviations, $\sigma_{g}$, uniform prior distributions on the interval 0 to 5,

\[
\tag{5}
p(\sigma_g) = 
\begin{cases}
    1/5 & \text{if } 0 \leq \sigma_g \leq 5 \\
    0   & \text{otherwise}
\end{cases}.
\]

These prior distributions were chosen based on preliminary analyses to allow adequate exploration of the parameter space while ensuring that the Markov chain Monte Carlo (MCMC) algorithm never halted by getting stuck in regions of negligible likelihood. These prior distributions should be viewed as weakly informative in the parlance of Bayesian analysis [@Lemoine2019]. 
The posterior distributions of the group-level averages, $p(\overline{\theta_g}|N_{AB})$, are proportional to the product of two likelihood distributions (Eqs. 1 and 3) and two prior distributions (Eqs. 4 and 5),

\[
\tag{6}
p(\overline{\theta_g}|N_{AB}) \propto p(N_{AB}|e^{\overline{\theta_g}+\Delta\theta})
p(\Delta\theta|\sigma_g) p(\overline{\theta_g}) p(\sigma_g).
\]

This expression is a hierarchical version of Bayes formula because the likelihood is the product of two distributions, $p(N_{AB}|e^{\overline{\theta_g}+\Delta\theta}) p(\Delta\theta|\sigma_g)$, rather than being a single distribution. Consequently, the model pools information for individuals pairs to estimate the within-group standard deviations, $\sigma_g$, and then draws values of $\Delta\theta$ from a normal distribution with a mean of 0 and a standard deviation of $\sigma_g$ to yield the pair-level estimates, $\theta_i$ (Eq. 2). Data pooling, in turn, results in shrinkage of pair-level estimates, $\theta_i$, toward the population-level means, $\overline{\theta_g}$.


A hierarchical approach is desirable here because pairs yielding weaker evidence (as indexed by smaller likelihoods) exhibit greater shrinkage, and pairs yielding stronger evidence are assigned greater weight in calculating the group-level averages, $\overline{\theta_g}$. Thus, this model fitting procedure accounts for the fact that species pairs with few records of occurrence, by themselves, yield little information on the value of $\theta$, and hence $\psi$. For example, if species A and B are both present at 5 of 25 sites, but there are no sites of co-occurrence (i.e. $N_{AB} = 0$, $N_{A^{+}} = 5$, $N_{A^{-}} = 20$, $N_{B+} = 5$), the maximum likelihood estimate is $\psi = 0$, implying complete segregation of species A and B. However, in this example, evidence of segregation is actually rather weak, as evidenced by the fact that maximum likelihood calculations (based on Eq. 1) yield a 95% confidence interval for $\psi$ that actually encompasses strong?? repulsion to strong attraction (0 to 6.1).


We estimate posterior distributions using two distinct approaches. First, we use hierarchical Bayesian MCMC methods to analyse full datasets, $N_{AB}$, comprised of all $S (S - 1)/2$ pairwise counts of co-occurrence for each of the two habitat types, where $S$ is the number of species used to calculate the set of pairwise estimates. This approach has the advantage of making use of all available data. However, it also has the disadvantage of not allowing for formal statistical evaluation because a single occupancy distribution for each species is used to generate $(S - 1)/2$ pairwise estimates of co-occurrence, resulting in non-independence.


Second, to address the issue of non-independence, we use Bayesian bootstrap aggregation [@Huggins2020], to obtain "bagged" estimates of posterior distributions, $p^{*}(\overline{\theta_g}|N_{AB}^{*}$), for independent samples, $N_{AB}^{*}$. Our approach entails the following: (\textit{i}) construct $D$ datasets comprised of independent co-occurrence counts ($N_{AB}^{*}$: $N_{AB}^{(1)}$, $N_{AB}^{(2)}$, ... $N_{AB}^{(D)}$); (\textit{ii}) sample the posterior distributions of the model parameters for each independent sample using Bayesian MCMC; (\textit{iii}) average over the $D$ posterior distributions obtained in (\textit{ii}) to approximate the posterior distribution for independent samples:

\[
\tag{7}
p^{*}(\overline{\theta_g}|N_{AB}^*) \approx \frac{1}{D} {\sum_{n = 1}^{D} p(\overline{\theta_g}|N_{AB}^{(n)})}
\]

For (\textit{i}), we first randomly permute a vector of species indices $1,...,S$, and then pair adjacent entries in the permuted vector, leaving $\lfloor S/2 \rfloor$ independent pairs, where $\lfloor x \rfloor$ is the largest integer $\leq x$. There are $S!/2^{\lfloor S/2 \rfloor}/\lfloor S/2 \rfloor!$ distinct independent datasets that can be constructed in this way because there are $S!$ ways permute the vector of $S$ species, $2^{\lfloor S/2 \rfloor}$ distinct ways to swap the ordering of species A and B within pairs, and $\lfloor S/2 \rfloor!$ ways to order the pairs. The number of distinct independent samples is far too large to consider them all. For example, with 100 species, there exist $100!/2^{50}/50! \approx 2.7\times 10^{78}$ distinct independent datasets of 50 pairs that can be constructed. We therefore approximated the bagged posterior distribution using $D$ = 100, as is standard [@Huggins2020].

For both types of analyses described above, we estimate the posterior distributions using the R package rjags, which provides an interface from R to the JAGS library for Bayesian data analysis [@Lunn2009; @Plummer2006]. To approximate the posterior distributions, we generated 3 MCMC chains of $5^5$ steps each in JAGS, including a burn-in period of 10000 steps, and a thinning interval of 100. We then inspected MCMC plots for the 3 chains and calculated Gelman Rubin statistics, $\hat{R}$, to ensure convergence [@Gelman2014]. For step (\textit{iii}) of our second analysis, we simply concatenated the posterior samples obtained from the $D=100$ trials because each MCMC trial yields a posterior sample of 
identical size, and all samples are given equal weight when calculating the bagged posterior distributions (Eq. 7).

```{r, echo = FALSE}
# calculate number of species per guild (in the functional analysis)
s <- spp %>% filter(rownames(.) %in% unlist(map(PAn, rownames))) %>% group_by(life.form, guild) %>% dplyr::summarise(count = length(guild))

```

In this framework, a change in the posterior distributions of group omegas and sigmas can exhibit several patterns, which point to various biological processes. Competing pairs can display lower omegas than non-competing pairs, which means that diffuse competition tends to lead to exclusion. This is the case even if all omegas are positive, as exclusion does not necessarily present as a negative association. Conversely, a higher omega in competing pairs confirms that similarity in resource use leads to co-occurrence without typically causing exclusion, suggesting that competing pairs are partitioning or sharing adequate resources. If the omegas of competing and non-competing pairs do not differ significantly, but the sigma (variance) of competing pairs is higher, this points to stronger spatial patterns (both coexistence and exclusion) in pairs that compete for food, and vice versa. If there is no difference in omega or sigma between the two groups, then there is insufficient evidence to support the hypothesis that diffuse competition is influencing the structure of the assemblage at the landscape scale.

The omega and sigma of competing and non-competing pairs might also differ between altered and unaltered sites in our example. If the relationship between competing and non-competing pairs is different, this means that alteration has changed the outcome of biotic interactions, e.g., by promoting partitioning over exclusion or vice versa. More subtly, the extent of disparity between the two groups could differ without qualitatively changing the relationship, which indicates that naturally existing mechanisms have been enhanced or dampened. On the other hand, omega or sigma could simply be higher or lower across the board, suggesting that alteration leads to changing community structure that does not necessarily involve the rewiring of competitive interations.  


*__The effect of turnover.__*
Overall changes observed in association patterns could result from pairs changing their behaviour or the intensity of their competition in altered vs. unaltered habitats. Conversely, a shift in species composition (e.g. the removal or introduction of interaction partners), and therefore the presence of a different suite of pairs, could also cause disparities between the two habitat types. To investigate these possibilities, we ran the models with all species and then only species occurring in both habitat types. 


#### Results and Discussion 
```{r, echo = FALSE}
# richness - p values
pcj1 <- cJ1 %>% group_by(taxon) %>% summarise(w=wilcox.test(richness~status, paired=FALSE)$p.value) %>% ungroup() %>% select(w)

pcha <- chao %>% group_by(taxon) %>% summarise(w=wilcox.test(richness~status, paired=FALSE)$p.value) %>% ungroup() %>% select(w)

# beta-diversity
beta <- beta.types(PAn, unalt_sites)
  # p-values
pbeta <- beta %>% filter(unalt.pair != "Unaltered-Altered") %>% mutate(similarity = qnorm(Z.Score)) %>% split(.$taxon) %>% purrr::map(~wilcox.test(data = ., similarity~unalt.pair, paired=FALSE))

# occupancy calculations
occ <- map(PAn[taxa], t)  %>% map(data.frame) %>% map(~split(., rownames(.) %in% unalt_sites)) %>% 
  map(map, ~colSums(.)/nrow(.)) %>% map(map, data.frame) %>% map(~merge(.[[1]], .[[2]], by = 0)) %>% map(setNames, c("name", "altered", "unaltered")) %>% bind_rows(.id = "taxon")
docc <- occ %>% group_by(taxon) %>% summarise(
  declined = length(which(altered < unaltered)), 
  increased = length(which(unaltered < altered)), 
  extirpated = length(which(altered == 0 & unaltered > 0)), 
  appeared = length(which(altered > 0 & unaltered == 0)), 
  shared = length(which(altered > 0 & unaltered > 0)), 
  total = length(altered))
```


*__Richness and beta diversity.__* Based on Chao 1 and cJ1, altered sites have lower alpha richness to unaltered sites, however the difference is not significant (cJ1 p-values  = `r round(pcj1[1,],3)` and `r round(pcj1[2,],3)`, and Chao 1 p-values  = `r round(pcha[1,], 3)` and `r round(pcha[2,], 3)` for bats and birds, respectively). Average richness for bat samples in unaltered habitats was 
`r cJ1 %>% group_by(taxon, status) %>% summarise(mean = mean(richness)) %>% {.[.$taxon == "bat" & .$status == "Unaltered",3]} %>% round(2)` (cJ1) and 
`r chao %>% group_by(taxon, status) %>% summarise(mean = mean(richness, na.rm = T)) %>% {.[.$taxon == "bat" & .$status == "Unaltered",3]} %>% round(2)` (Chao1), and
`r cJ1 %>% group_by(taxon, status) %>% summarise(mean = mean(richness)) %>% {.[.$taxon == "bat" & .$status == "Altered",3]} %>% round(2)` (cJ1) and 
`r chao %>% group_by(taxon, status) %>% summarise(mean = mean(richness, na.rm = T)) %>% {.[.$taxon == "bat" & .$status == "Altered",3]} %>% round(2)` (Chao 1) in altered habitats. Average richness for bird samples  was 
`r cJ1 %>% group_by(taxon, status) %>% summarise(mean = mean(richness)) %>% {.[.$taxon == "bird" & .$status == "Unaltered",3]} %>% round(2)` (cJ1) and 
`r chao %>% group_by(taxon, status) %>% summarise(mean = mean(richness, na.rm = T)) %>% {.[.$taxon == "bird" & .$status == "Unaltered",3]} %>% round(2)` (Chao 1) in unaltered habitats, and 
`r cJ1 %>% group_by(taxon, status) %>% summarise(mean = mean(richness)) %>% {.[.$taxon == "bird" & .$status == "Altered",3]} %>% round(2)` (cJ1) and 
`r chao %>% group_by(taxon, status) %>% summarise(mean = mean(richness, na.rm = T)) %>% {.[.$taxon == "bird" & .$status == "Altered",3]} %>% round(2)` (Chao 1) in altered habitats. 
For bats and birds, respectively, 
`r round(docc[1,3] / docc[1,7]*100)`% and 
`r round(docc[2,3] / docc[2,7]*100)`% of species increased their occupancies in altered habitats, and of these, 
`r round(docc[1,5] / docc[1,3]*100)`% and 
`r round(docc[2,5] / docc[2,3]*100)`%, respectively, appeared only in altered habitats. Overall, however, 
`r round(docc[1,2] / docc[1,7]*100)`% of all bat and 
`r round(docc[2,2] / docc[2,7]*100)`% of all bird species decreased their occupancies in altered habitats, and of these, 
`r round(docc[1,4] / docc[1,2]*100)`% and 
`r round(docc[2,4] / docc[2,2]*100)`% disappeared altogether. Relatively many species were sampled only in altered habitats (13% of bats and 27% of birds). These species may exist in unaltered habitats. However, as we took measures to standardise sampling between altered and unaltered sites, these species likely represent a group whose abundances have been substantially increased in altered sites. The converse may be true for some species that were only sampled in unaltered sites.

Altered sites had the lowest beta diversity in bats, decreasing significantly from unaltered sites  (Wilcoxon rank sum test, p _bat_ = `r round(pbeta[[1]]$p.value, 3)`). Birds also exhibited higher beta diversity in unaltered sites, but the difference was not significant (p _bird_ = `r round(pbeta[[2]]$p.value, 3)`). Bats had an order of magnitude higher similarity than birds overall, indicating that spatial turnover in birds is much higher and suggesting the two taxa have very different community structures. This is unsurprising because Neotropical bats frequently have geographic ranges spanning almost the entire realm, unlike Neotropical birds.

```{r, echo = FALSE}

# composition PCoA

dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "sample.no"))

# Significance test on centroids
w <- numeric()
for(j in 1:length(ord)){
  i <- ord[[j]]
  d <- numeric()
  for(r in 1:1000){
    i$rand = round(runif(n = nrow(i), min = 0, max = 1), 0)
    cent <- i %>% group_by(rand) %>% dplyr::summarise(x = mean(X1), y = mean(X2))
    d[r] <- spDists(as.matrix(cent[,c(2:3)]))[2]
  }
  cent <- i %>% group_by(status) %>% dplyr::summarise(x = mean(X1), y = mean(X2))
  obs <- spDists(as.matrix(cent[,c(2:3)]))[2]
  w[j] <- wilcox.test(d, obs)$p.value
}
w <- round(w, 3)
```

Figure 1 depicts the altered and unaltered sites plotted by the first two principal coordinates. The permutation tests detected no significant compositional differences between altered and unaltered sites (p-values = `r w[1]` for bats and `r w[2]` for birds). This indicates that the considerable faunal turnover observed across altered and unaltered sites is not atypical, even across sites in the same category.

In summary, unaltered and altered habitats did not differ significantly in species richness, nor did we find differences in composition between unaltered and altered habitats. This suggests that biogeographic location is more important in determining compositional turnover than habitat alteration. However, beta diversity was lower in altered habitats for bats, similar to the results of Kay _et al._ [-@Kay2017] for lizards in Australia. This paper therefore joins a long parade of studies showing that habitat alteration can cause homogenisation without reducing richness [@Toth2014; @Dornelas2014; @McKinney2006; @McKinney1999; @Trentanovi2013; @Knop2015]. We found this pattern even though our altered sites encompassed a wide variety of habitat types. Such homogenisation is typically driven by the expansion of a few cosmopolitan species while rare and sensitive species are extirpated [@McKinney1999; @Toth2014]. 

*__Biotic interactions.__*
Competition for food affected the co-occurrence structure of both bats and birds, as demonstrated by the offset between competing and non-competing pairs (yellow vs blue ovals, Figs. 2-5). Although competition did not always affect all aspects of associations, all of the subsets of pairs we examined exhibited significant differences between competing and non-competing pairs in at least one measured variable. 

Overall, in unaltered habitats, both bats and birds had weaker aggregations within competing pairs than non-competing pairs, but this pattern was reversed in altered sites (Fig. 2A and C). There was no effect of competition on bat segregations in unaltered habitats, but an effect appeared under altered condition, which competing pairs segregating more strongly than non-competing pairs (Fig. 2B). In birds, competing pairs segregate more strongly than non-competing pairs in unaltered habitats, but more weakly in altered habitats (Fig. 2D).   

This is the result we expect under natual conditions if competition is occurring in some species pairs that compete for food. Weaker aggregations can indicate that these pairs are in the process of excluding one another (co-occurring in some sites but not in others, perhaps interacting along an overlapping front), while the lack of stronger segregations suggest that exclusion has not occurred. In altered habitats, however, both aggregations and segregations of competing pairs were stronger on average than those of non-competing pairs, meaning habitat alteration reversed the effect of food competition on species aggregations (Fig. 2A), while displaying a novel effect for segregations. This suggests that under altered conditions, competing bat species either aggregate or exclude one another depending on the identity of the pair, forming clusters. 

*__The effect of turnover.__* 
Our analysis of compositional turnover showed that all types of pairs are affected by habitat alteration, although pairs with at least one unique species showed greater differences on average than shared-shared species pairs (Fig. 3). This confirms both hypotheses, namely, that food competition between species pairs occurring in both habitat types has a different outcome under altered conditions and that species occurring exclusively in altered habitats play a different structural role in the community than species occurring exclusively in unaltered habitats. By and large, shared-shared pairs in both bats and birds reflect the more general patterns described in the previous section (Fig. 3A-D). 
When shared species associate with species unique to unaltered and altered habitats, respectively, the general pattern is magified. In other words, competing species pairs consisting of one shared species and one species unique to unaltered habitats show strikingly stronger segregations and weaker aggregations than non-competing pairs (Fig. 3E-H). The only exception to this is segregations in birds, which do not indicate an influence of food competition (Fig 3H). This indicates that interaction between these shared-unique species pairs results in more stringent competition, and starker exclusion, than competition between two shared species in unaltered habitats. Conversely, a shared species competing with a species unique to altered sites results in strikingly stronger aggregations than non-competing pairs instead (Fig. 3E & G). 
Pairs consisting of unique species exhibit a different pattern altogether. In unaltered habitats, competing unique species pairs are consistently more strongly aggregated than non-competing pairs, and food competition appears to have no effect on segregations. This suggests a  nested nature of species distribution in pristine habitats, wherein a few hyper-diverse sites, or hotspots, contain a disproportionate number of unique endemic species as well as more common species. There is no evidence of competitive exclusion in these pairs. On the other hand, competing pairs unique to altered sites show no effect of competition on aggregations, instead exhibiting a heightened tendency to segregate compared to non-competing pairs. This suggests that among competing unique pairs only, alteration may promote exclusion, contrary to the overall pattern in the data.

*__The effect of disturbance preference in bats.__*
Several striking patterns emerged from our analysis of disturbance preferences in bats. The most conspicuous result is that pairs of synanthropic species (species that preferentially inhabit altered habitats) never aggregate in unaltered habitats (Fig. 4A). All pairs are segregated, and food competition has little effect on these segregations (Fig. 4G). This pattern is consistent with the notion that synanthropes are simply rare in unaltered habitats and reinforces the idea that habitat alteration engineers niches that are populated by novel, even artificial, communities. In altered habitats, food competition also appears to have little effect on synathropic pairs, although segregations are marginally, if significantly, stronger than in non-competing pairs, suggesting some instances of exclusion.
Competing cosmopolitan pairs (weak or no disurbance preference) have weaker aggregations when interacting with synanthropes or other cosmopolitan species in unaltered habitats. The latter group also exhibits stronger segregations, suggesting that diffuse competitive exclusion is in play in the absence of alteration. These patterns largely disappear in altered sites, suggesting that food competition is a much more important factor in unaltered sites for these pairs.
Restricted species (species that preferentially inhabit unaltered habitats) show another prominent pattern. There is very little offset between competing and non-competing pairs on the unaltered axis, suggesting minimal effect of food competition. Exclusion in competing pairs is only weakly evident (Fig. 4D-F, J-L), and is only significant when competition between a restricted species and a cosmopolitan species results in segregation. In stark contrast, food competition strongly influences community structure in altered habitats, causing stronger segregations with all types of interaction partners. Futhermore, the relationship between competing and non-competing aggregated pairs undergoes a reversal depending on the disturbance preference of the interaction partner. Competing pairs exhibit much stronger aggregations than non-competing pairs when a restricted species is paired with a synanthrope (Fig. 4D), there is no effect when a restricted species is paired with a cosmopolitan species (Fig. 4E), and aggregations are drastically weakened when both species are restricted (Fig. 4F). 
This last pattern is doubly conspicuous because it reverses expectations. If restricted species are simply rare in altered habitats, we would expect aggregations (if any exist) to be strong (as they are unlikely to occur by chance) and segregations to be weak, with no evidence of food competition effect, much like the observed inverse pattern in synanthropic pairs. This is also the reverse of the pattern we would expect to see if co-occurrence is being influenced by the previously recorded geographic structure of bat dietary guilds [@Alroy2018??]. Therefore, it appears that restricted-restricted bat pairs experience fierce competition when subjected to unfavourable conditions, a pattern which is the reverse of the overall trend in the data. (Is it possible that this characteristic has a role in determining which species end up being restricted?!)

*__The effect of disturbance preference in birds.__*
Disturbance preference had less influence on co-occurrence patterns in birds than bats, but there are nonetheless some overarching patterns (Fig. 5). In pairs comprised of synanthropic and cosmopolitan species, food competition appears to play a role in community structure in unaltered habitats, and the effects are, by and large, eradicated under altered conditions (the sole exception being Fig. 5C, where aggregations are stronger for competing pairs in altered sites). This loosely mirrors the results for bats. For pairs involving at least one restricted species, however, food competition has significant effects in altered habitats, while unaltered habitats exhibit varying degrees of competitive influence (Fig. 5 D-F, J-L). Once again, contrary to expectation, pairs in which both species are restricted show all the hallmarks of competitive exclusion.

*__Altered site types.__*
 Strong aggregations form when species occur together at a relatively small number of sites and are both absent from other sites. Vampire bats have been shown to feed preferentially on the blood of livestock [@Delpietro1992; @Bobrowiec2015], suggesting they co-occur in farms and ranches [@Becker2018]. To help understand our results, we calculated the representation of dietary guilds at various different types of altered sites (Fig. S6). Although there are only a few species of sanguinivore bats in our data, carnivore-insectivorous (22 species) and frugivorous bats (69 species) co-occur disproportionately in plantations. Plantations and forest fragments support the highest proportion of insectivore bats (80 species), and nectarivores (21 species) co-occur disproportionately in plantations and croplands (see supplement and Fig. S6). Non-competing pairs might segregate due to this spatial structure in food resources.

For birds, inhabited sites support a disproportionate ratio of birds with simple diets (nectarivores, frugivores, invertivores, and granivores), though the sample size is too small to derive any conclusions. Croplands are the main supporter of frugivore-granivore birds.

#### Conclusions 

 Our approach confirms that the effect of food competition on community structure can be detected using co-occurrence. There were marked differences in the community assembly of bats and birds as a result of habitat alteration, even though richness, turnover, and composition were not significantly different. Most competing pairs exhibit stronger aggregations in altered habitats and stronger segregation in unaltered habitats, suggesting that changes in community assembly are influenced by the spatial aggregation of food resources in altered sites. This implies that habitat alteration reduces competitive exclusion for food, particularly in bats. Unfortunately, this leads to reduced co-occurrence of functionally diverse pairs, as evidenced by the comparatively low co-occurrence of non-competing pairs. 
 
Species that have persisted in both types of sites have differing co-occurrence patterns, tending to exhibit a reduced role for food competition in altered sites. We did not evaluate the direct causes in this study, but changes in the spatial allocation of food resources, the creation of novel niches, changes in behaviour, and even genetic adaptations are all possible causes. Bats often share roosts with other species [@Swift1983] and have been shown to partition microhabitats and foraging times [@Aldridge1987; @Nicholls], which allows them to overlap broadly on the trophic dimension [@Kingston2000]. The changes associated with habitat alteration enhance this pattern, suggesting that certain properties of altered habitats actively aggregate particular diet guilds in space (e.g. to particular types of sites) or in time (e.g. seasonal fruiting of a frequently planted tree). Birds segregate much more often than bats in general, but competing pairs are nonetheless more aggregated in altered habitats, suggesting that a similar mechanism may be in play for both taxa. 

Species found uniquely in altered habitats play opposite structural roles than uniquely unaltered species do in their own domain. When competing with shared species, the former aggregate strongly and the latter aggregate weakly compared to non-competing pairs. When competing with one another, the former form marginally stronger segregations (no effect on aggregations) and the latter form very strong aggregations (no effect on segregations) with respect to non-competing pairs. Therefore, species compositional turnover plays a key role in changed patterns of competition in altered habitats.

The effect of competition on community structure was weaker in altered habitats than unaltered habitats for synanthropic and cosmopolitan species. When competition still had an effect, it tended to enhance aggregations rather than segregations, possibly because food resources are more reliable or abundant. This general trend drove the overall pattern in all pairs and agrees with the results for the broader group of shared-shared pairs. Species that prefer unaltered habitats, by contrast, have considerably stronger segregations in altered habitats when they compete with all types of interaction partners (relative to non-competing pairs), opposite to the overall patterns in all pairs. 

Although some differences did exist, the qualitative responses of competing bats and birds to habitat alteration had strong parallels. Overall, habitat alteration reduces or reverses the effect of competition on community structure in both bats and birds, with restricted species, whose ability to coexist is greatly compromised by disturbance, being of particular concern as altered habitats become increasingly widespread. 


#### Acknowledgements 
The authors would like to thank the members of the Macquarie University Paleobiology Lab, whose discussions helped facilitate the improvement of this manuscript. ABT was partially supported by an IMQRES scholarship at Macquarie University. 

### Figures

```{r echo=FALSE, include = T, fig.height=4, fig.width=9, ord}
# Ordination
dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "sample.no"))

#PCoA plot
  p1 <- ggplot(ord[[1]], aes(x = X1, y = X2, col = status)) + geom_point() + labs(x = "Comp1", y = "Comp2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("red", "dodgerblue"))
  
  p2 <- ggplot(ord[[2]], aes(x = X1, y = X2, col = status)) + geom_point() + labs(x = "Comp1", y = "Comp2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("red", "dodgerblue")) 
  
  p <- plot_grid(p1+theme(legend.position = "none"), 
                 p2+theme(legend.position = "none", 
                axis.title.y = element_blank()), ncol = 2, labels = c("A","B"), align = "lr")
  plot_grid(p, get_legend(p1), rel_widths = c(1, .2))

```

__Figure 1.__ Principal coordinates analysis of altered and unaltered sites for bats (A) and birds (B). There is no significant difference between the composition of altered and unaltered sites for either taxon, based on a permutation test of the site type centroids (p-values = `r w[1]` and `r w[2]` for bats and birds, respectively). 

```{r}
load("../Results/d2_d5_summary_tables_nr_ns_100r_relatedexcl.RData")
colors2 <- c("#34008C", "#E8AC00")

specs <- list(geom_hline(yintercept = 1, col = "gray"),
              geom_vline(xintercept = 1, col = "gray"),
              stat_ellipse(geom= "polygon", alpha = 0.3, aes(col = diet.match, fill = diet.match)),
              geom_point(aes(col = diet.match, fill = diet.match), size = 0.5), 
              geom_rug(aes(col = diet.match), alpha = 0.5),
              scale_colour_manual(values = colors2),
              scale_fill_manual(values = colors2),
              theme(legend.position = "none")
)
```


```{r, echo=FALSE, include = T, fig.height=4, fig.width=4}
# Agg/seg strength in bird and bat pairs

d <- d2.mag.all
# prepare data
obsexp <- obsDexp(d, split.var = "type", data.var = "avmag", Taxon_status, diet.match, pnz)
 #b <- bayesPairedTtest(obsexp, split.var = "diet.match", taxon, pnz)

anno <- obsexp %>% group_by(taxon, pnz) %>% 
  summarise() %>% ungroup() %>% 
  mutate(label = paste0("  ", LETTERS[1:4]), x = -Inf, y = Inf)

ggplot(obsexp, aes(x = unaltered, y = altered)) +
  specs + facet_wrap(pnz~taxon, scales = "free") +
  geom_text(data = anno, aes(x = x, y = y, label = label), vjust = "inward", hjust = "inward", fontface = 2)

```

__Figure 2.__ Strength of aggregations and segregations in bat and bird pairs. Each point represents the mean observed-to-expected ratio for one run of the randomisation. Yellow points are means for competing pairs and dark blue points are means for non-competing pairs in each run. 95% confidence ellipses are shown in the same colour scheme. Results for altered sites are plotted against unaltered results. Gray lines represent 1:1 observed-to-expected ratios, and are shown to help interpret differences between the two ovals. See Figure S2 for more information. 

```{r, echo=FALSE, include = T, fig.height=8, fig.width=8}
## Shared/Unique
d <- d2.mag.catp
obsexp <- obsDexp(d, split.var = "type", data.var = "avmag", Taxon_status, diet.match, cat.pair, pnz)

# bayesian paired t-test on competing and non-competing pairs
   # b <- bayesPairedTtest(obsexp, split.var = "diet.match", taxon, cat.pair, pnz)
 
anno <- obsexp %>% group_by(cat.pair, taxon, pnz) %>% 
  summarise() %>% ungroup() %>% 
  mutate(label = paste0("  ", LETTERS[1:12]), x = -Inf, y = Inf)

ggplot(obsexp, aes(x = unaltered, y = altered)) + 
  specs+ facet_wrap(cat.pair+taxon~pnz, scales = "free") +
  geom_text(data = anno, aes(x = x, y = y, label = label), vjust = "inward", hjust = "inward", fontface = 2)

```

__Figure 3.__ Strength of aggregations and segregations comparing competing (yellow) and non-competing (dark blue) pairs when separated into shared-shared (A-D), shared-unique (E-H), and unique-unique (I-L) pairs. Results from altered sites are plotted against results from unaltered sites. Refer to Figure S2 and caption of Figure 2 for more information.

```{r, echo=FALSE, include = T, fig.height=7.5, fig.width=3, warning=FALSE}
# Synanthropic/Cosmopolitan/Restricted

d <- d5.mag.catp
obsexp <- obsDexp(d, split.var = "type", data.var = "avmag", Taxon_status, diet.match, cosmo.pair, pnz)
obsexp$cosmo.pair <- factor(obsexp$ cosmo.pair, levels = c("synan-synan", "cosmo-synan", "cosmo-cosmo", "restr-synan", "restr-cosmo", "restr-restr"))
 # bayesian paired t-test on competing and non-competing pairs
  # b <- bayesPairedTtest(obsexp, split.var = "diet.match", taxon, cosmo.pair, pnz)

anno <- obsexp %>% filter(taxon == "bat") %>% group_by(pnz, cosmo.pair) %>% 
  summarise() %>% ungroup() %>% 
  mutate(label = paste0("  ", LETTERS[1:12]), x = -Inf, y = Inf)

ggplot(obsexp %>% filter(taxon == "bat"), aes(x = unaltered, y = altered)) + 
  specs+ facet_wrap(cosmo.pair~pnz, scales = "free", ncol = 2, nrow = 6) +
  geom_text(data = anno, aes(x = x, y = y, label = label), vjust = "inward", hjust = "inward", fontface = 2)


```

__Figure 4.__ Strength of aggregations and segregations comparing competing (yellow) and non-competing (dark blue) bat pairs when separated into groups based on disturbance preference of species. Results from altered sites are plotted against results from unaltered sites. Refer to Figure S2 and caption of Figure 2 for more information.

```{r, echo=FALSE, include = T, fig.height=7.5, fig.width=3, warning=FALSE}
anno <- obsexp %>% filter(taxon == "bird") %>% group_by(pnz, cosmo.pair) %>% 
  summarise() %>% ungroup() %>% 
  mutate(label = paste0("  ", LETTERS[1:12]), x = -Inf, y = Inf)

ggplot(obsexp %>% filter(taxon == "bird"), aes(x = unaltered, y = altered)) + 
  specs+ facet_wrap(cosmo.pair~pnz, scales = "free", ncol = 2, nrow = 6)+
  geom_text(data = anno, aes(x = x, y = y, label = label), vjust = "inward", hjust = "inward", fontface = 2)

```

__Figure 5.__ Strength of aggregations and segregations comparing competing (yellow) and non-competing (dark blue) bird pairs when separated into groups based on disturbance preference of species. Results from altered sites are plotted against results from unaltered sites. Refer to Figure S2 and caption of Figure 2 for more information.


## References



