---
title: Competitive interactions are disrupted by habitat alteration in the Neotropics
output:
  pdf_document: 
    dev: png
  bookdown::word_document2: default
bibliography: References.bib
csl: nature.csl
---

Anikó B. Tóth, Andrew P. Allen, S. Kathleen Lyons, John Alroy

*This PDF file includes:* 

Materials and Methods

Supplementary Text

Figs. S1 to S3

Tables S1 to S2

References (38–55)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

inline_hook <- function(x) {
  if (is.numeric(x)) {
    format(x, digits = 2)
  } else x
}
knitr::knit_hooks$set(inline = inline_hook)

## Load helper functions
source('../Code/HelperFunctions.R')
## Load stan functions
source('../Code/stan.R')
## Prep raw data
source('../Code/Data_Prep.R')

library(tidyverse)
library(cowplot)
library(reshape2)
library(vegan)
library(flextable)
library(bayestestR)
library(lsa)
library(plotrix)
```

### Materials and Methods

_Data_ 

Abundance data were downloaded from the Ecological Register (http://ecoregister.org) on 7th June 2020 for mist-netted bats and birds from the Neotropics, defined as all areas in the western hemisphere south of the Tropic of Cancer (23.44˚N). We also downloaded Register metadata for the species and sites in the abundance tables, including coordinates and habitat alteration categories for sites and dietary guild assignments for species. Guild assignments were based on a large set of recently published papers. The Ecological Register provides methodologically consistent abundance data, not ad hoc occurrence data. Thus, sampling and reporting were standardized across the samples: all were mist-netted samples with consistent mesh size; harp net samples were excluded, and almost all nets were placed at ground level. Every sample consisted of a minimum of 20 individuals and the vast majority of sites had more than 50, with the median number of individuals per sample being `r median(colSums(PAn[["bat"]]))` for bats and `r median(colSums(PAn[["bird"]]))` for birds.  Samples were combined if they were less than 1 km from one another, sourced from the same study, and represented the same habitat and alteration category. The combined data set consisted of `r sum(PAn[["bat"]])` bat and `r sum(PAn[['bird']])` bird individuals constituting `r length(which(PAn[["bat"]] > 0))` bat and `r length(which(PAn[["bird"]] > 0))` bird records from `r ncol(PAn[['bat']])` and `r ncol(PAn[['bird']])` sites, respectively. Maps of the sites are plotted in Figure S1.

_Biogeographic correction_ 
```{r biogeocorr, include = FALSE}
#### Match Biogeography ####
# separate altered and intact sites
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites])) 
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites])) 

coords <- sitedat %>% select(taxon, p.sample, latitude, longitude) %>% distinct() %>% split(.$taxon)

# Remove intact or altered sites that are not near a site of the other type. This is done to ensure the two sets have similar biogoegraphical distributions.
coords1 <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2 <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])
keep <- map2(coords1, coords2, matchbiogeo) %>% map(unlist)

PAn <- map2(PAn, keep, function(x, y) return(x[,as.character(y)]))
PAn <- map(PAn, clean.empty, minrow = 1) # remove any species that now have no occurrences

# recalculate alt/unalt split from new PAn
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites]))
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites]))

# coords of sites we kept, for plotting later.
coords1.keep <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2.keep <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])

```

The geographic layout of sites can influence the results of any community ecology analysis, including co-occurrence analysis. For instance, if sites are clustered in two regions with no sites in between, there is a good chance that the co-occurrence structure of the assemblage will reflect this clustered layout by creating strong aggregations within the clusters and segregations between clusters. Ideally, we would like to have even sampling over our entire study area, but when this is not possible, we must at least make sure that any comparisons are not influenced by geographic sampling bias. One way to achieve this is to ensure that the sites used to calculate co-occurrences have roughly the same geographic coverage when comparing different types of sites.

In this paper, the study extent covers Central and South America south of the Tropic of Cancer. This includes large swathes of the Brazilian Amazon, where there are no samples in altered habitats. As a result, the intact sites cover more area than the altered sites, and the geographic coverage of the latter is nested within that of the former. To correct for these biases before calculating co-occurrences, we removed the intact sites covering areas that were not covered by altered sites using the following algorithm: (1) we measured the distance to the closest altered site for all intact sites and vice versa; (2) we found the altered site farthest from its closest intact site; (3) we removed all intact sites whose closest altered neighbors were farther than the distance in step 2. Note that this approach only works because the geographic coverages are nested; site types that are offset in space would require a different approach. The resulting dataset had `r sum(PAn[["bat"]])` occurrences of bats and `r sum(PAn[["bird"]])` occurrences of birds at `r ncol(PAn[["bat"]])` and `r ncol(PAn[["bird"]])` sites, respectively. Our original and final sites are plotted in Fig. S1.

_Richness_ 
```{r richness, message=FALSE, warning=FALSE, include=FALSE}
###### Richness - run using raw abundance data ####
# can only be run on raw data because it requires a singleton count of abundances.
rich <- list(cJ1 = map(PAn, map_dbl, cJ1rich) %>% map(~split(., names(.) %in% unalt_sites)) %>% # corrected 1st order jackknife
               map(map, cbind) %>% map(map, data.frame), 
             chao = map(PAn, map_dbl, chao1) %>% map(~split(., names(.) %in% unalt_sites)) %>%  # chao1
               map(map, cbind) %>% map(map, data.frame) , 
             fa = map(PAn, map_dbl, fisher.alpha) %>% map(~split(., names(.) %in% unalt_sites)) %>%  # fisher's alpha
               map(map, cbind) %>% map(map, data.frame)) %>% 
  map(map, bind_rows, .id = "status") %>% map(bind_rows, .id = "taxon") %>% bind_rows(.id = "metric") %>% 
  setNames(c("metric", "taxon", "status", "richness")) %>% 
  mutate(status = recode(status, `FALSE` = "Altered", `TRUE` = "Intact"))

# Summary
rsumm <- rich %>% group_by(metric, taxon, status) %>% summarise(mean.rich = mean(richness, na.rm = T), median.rich = median(richness, na.rm = T))
# Significance
rsig <- rich %>% group_by(metric, taxon) %>% summarise(p=wilcox.test(richness~status, paired=FALSE)$p.value, 
                                               W=wilcox.test(richness~status, paired=FALSE)$statistic)


```

We estimated richness for bats and birds in altered and intact sites using three diversity metrics: a corrected first-order jackknife (cJ1) [@Alroy2020], the canonical Chao1 [@Chao1984], and Fisher’s alpha, which is robust to sample size, widely used, and well-grounded in theory [@Schulte2005]. Each metric was applied to the raw abundance data for all sites. The status of each site was based on the ‘altered.habitat’ field in the site metadata, which is empty if the site is intact and filled with a land use category if the site is altered. Altered sites included cropland, disturbed forest, fragment, pasture, plantation, secondary forest, inhabited area, and combined categories [@Alroy2017], and none of them had sufficient sites to be evaluated individually. We compared richness estimates between altered and intact sites for both taxa using a Wilcoxon rank-sum test.

_Beta diversity and species composition_ 
```{r betacomp, message=FALSE, warning=FALSE, include=FALSE}

#### Beta diversity and composition analyses #####
output <- data.frame()
input <- list(abundance = PAn, binary = tobinary(PAn))

for(d in c("binary", "abundance")){
  for(metric in c("Jaccard", "Ochiai")){

PA <- input[[d]]
if(metric == "Jaccard") dist <- map(PA, ~t(.)) %>% map(vegdist, method = "jaccard") else dist <- map(PA, as.matrix) %>% map(cosine) %>% map(as.dist, upper = F) %>% map(~return(1-.))

if(metric == "Ochiai" && d == "abundance") metric <- "Cosine similarity"
# beta diversity using betadisper
beta <- map2(dist, PA, function(x, y) betadisper(x, group = colnames(y) %in% unalt_sites) %>% anova)

# compositional change using adonis
data <- map(PA, ~t(.x) %>% data.frame() %>% rownames_to_column("p.sample") %>% mutate(ID = colnames(.x) %in% unalt_sites))
comp1 <- map2(dist, data, function(x, y) adonis(x~ID, data = y, permutations = 10000))

# compositional change using canonical correspondence analysis
dat <- map(data, select, contains("_"))
comp2 <- map2(dat, data, function(x, y) cca(x~ID, data = y) %>% anova.cca)

bdc <- list(beta, map(comp1, ~.x$aov.tab), comp2)
output <- rbind(output, data.frame(analysis = c("beta dispersion", "adonis", "cca"), 
                             data = d, 
                             metric = c(metric, metric, "NA"), 
                             map_df(bdc, map_df, ~.x$P[1]) %>% setNames(c("p bat", "p bird")), 
                             map_df(bdc, map_df, ~.x$F[1]) %>% setNames(c("F bat", "F bird"))))
  }
    
}



```

We compared the beta diversity and species composition of altered and intact sites using a suite of classical analyses. A common way to estimate beta diversity differences between two groups of samples is a multivariate dispersion test [@Anderson2006], which can be implemented using the *betadisper* function in the R package *vegan* [@Oksanen2008]. We also compared the composition of altered and intact sites for both taxa with permutational multivariate ANOVA using *adonis* [@Anderson2001] and canonical correspondence analysis using *cca* [@Legendre2012], also from the *vegan* package. Multivariate dispersion and permutational ANOVA require a distance metric, and the choice of distance metrics may influence the results. We therefore ran these analyses using the relatively ubiquitous Jaccard index [@Hubalek1982] and again with the Ochiai index [@Hubalek1982]. Beta diversity and composition can be compared based on abundance data or occurrence data, noting that the Ochiai index is equivalent to cosine similarity when run on abundance data. We ran our analyses with both data types, but we focus on the results obtained with occurrence data, as these are much more common and easily obtained. However, we note those cases where results from abundance data differ quantitatively.

_Biotic interactions_ 
```{r message=FALSE, warning=FALSE, include=FALSE}
PAnb <- tobinary(PAn)
tables <- PAnb %>% map(~t(.)) %>% map(as.data.frame) %>% map(~split(., f = rownames(.) %in% unalt_sites)) %>% 
  map(map, ~t(.)) %>% map(map, clean.empty) %>% purrr::map(setNames, c("altered", "unaltered"))

original <- map(tables, map, ~expand.grid(Sp1 = rownames(.x), Sp2 = rownames(.x))) %>% map(map, diet_cat, spp, related = TRUE) %>% map(bind_rows, .id = "status") %>% bind_rows(.id = "taxon") %>% na.omit() %>% filter(Sp1 != Sp2) %>% select(taxon, Sp1, Sp2, diet.match) %>% distinct() %>% group_by(taxon) %>% summarise(related = length(which(diet.match == "Related")), pct_related = 100*related/n())

#no-turnover
 #shared <- tables %>% map(~.x %>% map(rownames) %>% reduce(match_val))
 #tables <- map2(tables, shared, function(x, y) map(x, function(z) return(z[y,])))

 #noturnover <- map(tables, map, ~expand.grid(Sp1 = rownames(.x), Sp2 = rownames(.x))) %>% map(map, diet_cat, spp, related = TRUE) %>%  map(bind_rows, .id = "status") %>% bind_rows(.id = "taxon") %>% na.omit() %>% filter(Sp1 != Sp2) %>% select(taxon, Sp1, Sp2, diet.match)   %>% distinct() %>% group_by(taxon) %>% summarise(related = length(which(diet.match == "Related")), pct_related = related/n())

```

We used dietary information for bats and birds from the Ecological Register to analyse the effects of food competition on patterns of co-occurrence. We selected all pairs with complete dietary information and classified them as competing or non-competing. Competing pairs shared primary and secondary sources of food, but we allowed the order of these to be reversed (e.g., a frugivore-nectarivore and a nectarivore-frugivore would be equated). Non-competing pairs had no mutual food sources. Pairs with mutual and nonmutual food sources (e.g., a frugivore and an insectivore-frugivore) may experience a complex mix of effects and were therefore removed from the analysis for simplicity (`r original %>% filter(taxon == "bat") %>% pull(related)` or `r original %>% filter(taxon == "bat") %>% pull(pct_related)`% of bat pairs and `r original %>% filter(taxon == "bird") %>% pull(related)` or `r original %>% filter(taxon == "bird") %>% pull(pct_related)`% of bird pairs fell into this category). Although species in the same dietary guilds do not necessarily compete, pairs with high dietary overlap have a nonzero probability of experiencing competition for food, while pairs in differing guilds have no chance of competing for food if the categorizations are accurate. Therefore, the co-occurrence patterns of competing and non-competing groups should be different if food competition plays a role in structuring a given assemblage.

This framework also allows the modeler to quantify the effect of various external factors and ecological contexts on the spatial outcomes of biotic interactions. In this paper, we compare assemblages from altered and intact habitats to evaluate the effect of habitat alteration on the outcome of food competition. However, the samples could be split up along other lines instead, such as along environmental gradients, temporally before and after disturbances, etc.

Recent studies repeatedly point out that nonrandom co-occurrence should not be construed as evidence for interaction between species pairs [@Freilich2018;@Thurman2019; @Blanchet2020]. In our framework, we have independent evidence of potential interactions (diet) and examine the relationship between putative interactions and co-occurrence, using non-interacting pairs as a control. Controls are rarely used in co-occurrence research, with most studies preferring to discard pairs with associations that are not seen as significant (e.g. [@Araujo2011; @Lyons2016]). Research using such an approach is exactly what is needed to understand under what circumstances, and to what extent, interactions lead to nonrandom spatial associations, what factors affect the relationships, and how they change under shifting conditions.

\newcommand{\vect}[1]{\boldsymbol{#1}}
\renewcommand{\thefigure}{S\arabic{figure}}

_Fisher's noncentral hypergeometric distribution_
We assume that the probability that the number of sites of co-occurrence,
$N_{AB}$, for a given pair of taxa, A and B, adheres to Fisher's noncentral
hypergeometric distribution (NHD) [@Fog2008],

\begin{equation}
\tag{S1.a}
f(N_{AB} \mid \theta) =
\frac{\binom{N_{A}}{N_{AB}} \binom{N_{(A)}}{N_{B} - N_{AB}} \exp(N_{AB} \theta)}
{P_0(\theta)}
\end{equation}

where $N_{A}$ and $N_{(A)}$ are the respective numbers of sites with taxon A
present and absent, $N_{A} + N_{(A)}$ is the total number of sites, $N_{B}$ is
the number of sites with taxon B present, and $\binom{n}{k} = n!/k!/(n-k)!$ is a
binomial coefficient. The denominator of this probability mass function (pmf),
$P_0(\theta)$, is the numerator summed over the range of possible values for the
number of co-occurring sites [@McCullagh1952], 

\begin{equation}
\tag{S1.b}
P_k(\theta) = \sum_{n = \max(0,N_{B}-N_{(A)})}^{\min(N_{A},N_{B})} 
\binom{N_{A}}{n} \binom{N_{(A)}}{N_{B} - n} 
\exp(n \theta) n^k
\end{equation}

This pmf has one fitted parameter,

\begin{equation}
\tag{S1.c}
\theta = \log(\psi) =\log \left( \frac{\pi_{B|A}}{\pi_{B|(A)}} \right) = 
\log \left( \frac{p_{B|A} (1 - p_{B|(A)})}{(1-p_{B|A}) p_{B|(A)}} \right)
\end{equation}

which is the logarithm of the odds ratio $\psi = \pi_{B|A}/\pi_{B|(A)}$, where
$\pi_{B|A}$ and $p_{B|A}$ are the respective odds and probability that taxon B
occupies a site given that taxon A is present, and $\pi_{B|(A)}$ and $p_{B|(A)}$
are the respective odds and probability that taxon B occupies a site given that
taxon A is absent. The values of $\theta$ and the probability of $N_{AB}$
co-occurrences, $f(N_{AB}|\theta)$, are independent of which member of the taxon
pair is arbitrarily chosen as taxon A. The NHD is typically parameterized using
$\psi$ rather than $\theta$, but $\psi$ has a lower bound of 0, making it less
convenient for model fitting and biological interpretation.


The NHD parameter $\theta$ directly quantifies spatial patterns of co-occurrence
for a taxon pair. If $\theta < 0$, taxon B is less likely to occur at a site if
taxon A is present, as we would expect if there was competitive exclusion or
differences in habitat preferences between the two taxa. If $\theta>0$, taxon B
is more likely to occur at a site if taxon A is present, as we would expect if
the taxa shared habitat preferences and did not competitively exclude each
other. Finally, if $\theta = 0$, the two taxon are independently distributed.
When $\theta = 0$, the NHD reduces to the standard (i.e.\ centered)
hypergeometric distribution, which has frequently been used to evaluate patterns
of co-occurrence [@Arita2016].
To our knowledge, our study is the first to use the noncentral form of the
hypergeometric distribution for this purpose. 

In light of the rich array of methodologies already available for quantifying species associations, the use of a novel method must be justified. First, while it is mathematically related to existing null-model-based co-occurrence metrics which have been shown to be relatively reliable (e.g. mid-P variant of Fisher's Exact Test, see [@Keil2021]), $\theta$ is a direct parametrization of the effect of one species on the occurrence of another (Equation S1.c). Rather than using the observed data to accept or reject the significance of a co-occurrence pattern as older methods do, it attempts to estimate the underlying effect that produces the observed data (similar to the difference between a p-value and a correlation coefficient). For this reason, existing methods are not commensurate with our approach, and we do not attempt to draw any direct comparisons. 

This parametric approach represents a significant conceptual advance for several reasons. First, it theoretically allows for the assocation of species pairs to vary even if their observed occurrences are the same, as we should expect in real life. This distinction becomes important when a Bayesian model is used to estimate the value of the parameter. Second, it is strongly biologically interpretable, quantifying specifically the effect of one species’ occupancy on another’s while being easily programmable in Stan. Third, $\theta$ is not sensitive to the number of sites in the study assemblage, making formal comparisons and meta-analyses possible down the track. With these improvements, our analysis is sufficiently different from existing co-occurrence approaches that a direct comparison between the two is not only not sensible but also not needed to justify the new analysis. Should the comparison between the two approaches yield different results, we should be more confident in the results of the present approach above the results from null model analyses. However, we perform some analyses to demonstrate the behaviour of the NHD parameter.


To explore properties of the NHD, we calculate how the distribution of $N_{AB}$
values varies with $\theta$ (ranging from -2 to 2) for occurrence values that
span the majority of taxa in our datasets ($N_{A}$ and $N_{B}$ ranging from 2 to
8; Figure S2). The total number of sites for these calculations was set at 50, which is
similar to the values for the bird and bat datasets we analyzed (47 and 53,
respectively). Besides demonstrating that increases in $\theta$ correspond to higher probabilities of more frequent co-occurrences, as we would expect from a reliable metric, two noteworthy observations arise from these calculations. First,
if occurrences of the two taxa are low relative to the total number of sites
($N_{A}=N_{B}=2$), $N_{AB}=0$ is the most likely outcome even when $\theta=2$,
which corresponds > 7-fold higher odds of taxon B occurring at a site if taxon A
is present (solid red line in Figure S2). Second, as the numbers of sites
occupied by the two taxa increases, the distribution of $N_{AB}$ values exhibits
a greater shift in mode with changes in $\theta$: when $N_{A}=N_{B}=2$, the mode
for $N_{AB}$ remains at 0 for all three $\theta$ values (-2, 0, 2), whereas when
$N_{A}=N_{B}=8$, the mode shifts from 0 (at $\theta=-2$) to 4 (at $\theta=2$).


To understand the maximum likelihood of the NHD, we take logarithms of both sides of
Eq.\ (S1.a), and differentiate with respect to $\theta$, yielding an expression
for the \emph{score}, defined as the gradient of the log-likelihood in relation
to the parameter of interest (in this case $\theta$):

\begin{equation}
\tag{S2}
\frac{\partial \log f(N_{AB} \mid \theta)}{\partial \theta} = 
N_{AB} - P_1(\theta)/P_0(\theta)
\end{equation}

The quantity $P_1(\theta)/P_0(\theta)$ is the expectation (i.e.\ average) for
the NHD (Generalized linear models McCullagh, P. (Peter), 1952-; Nelder, John A.
2nd ed. London ; New York : Chapman and Hall 1989). Given that the maximum
likelihood for $\theta$, $\hat{\theta}$, is achieved when $\partial \log
f(N_{AB} \mid \theta) / \partial \theta = 0$, Eq.\ (S2) implies that
$\hat{\theta}$ corresponds to the NHD with a mean equal to the observed number
of co-occurrences, $N_{AB}$. (While there is no exact analytical expression for
this mean, accurate approximations have been derived [@Eisinga2011]. Consequently, the estimate of
$\hat{\theta}$ inferred from a single observation is biologically uninformative
whenever the observed number of co-occurring sites is equal to the minimum or
maximum possible values, i.e.\ $\hat{\theta}=-\infty$ if
$N_{AB}=\max(0,N_{B}-N_{(A)})$, and $\hat{\theta}=\infty$ if
$N_{AB}=\min(N_{A},N_{B})$.


Our capacity to infer $\theta$ from observations is fundamentally constrained by
the Fisher Information, which is the variance of the score [@Ly2017]. We
can characterize how the Fisher information varies with $\theta$,
$I(\theta)$, by combining Eqs.\ (S1) and (S2):

\begin{equation}
\tag{S3}
\begin{split}
I(\theta) & = E \left[
\left( \frac{\partial \log f(N_{AB} \mid \theta)}{\partial \theta} \right)^2 
\mid \theta \right] = \\
& = \sum_{n = \max(0,N_{B}-N_{(A)})}^{\min(N_{A},N_{B})} 
f(n \mid \theta) \left( \frac{\partial 
\log f(n \mid \theta)}{\partial \theta} \right)^2 =
\frac{P_2(\theta)}{P_0(\theta)} - 
\left( \frac{P_1(\theta)}{P_0(\theta)} \right)^2
\end{split}
\end{equation}

Remarkably, Eq.\ (S3) demonstrates that, when the NHD is parameterized using
$\theta$, the Fisher Information is
$P_2(\theta)/P_0(\theta)-(P_1(\theta)/P_0(\theta))^2$, which is also the
variance of the NHD [@McCullagh1952].
Parameterizing the NHD using the odds ratio $\psi$ would yield a different
expression for the Fisher Information.


Calculations performed using Eq.\ (S3) demonstrate that the Fisher Information
generally increases with the taxon occurrences $N_{A}$ and $N_{B}$ (Figure S3).
Loosely speaking, this result indicates that the distribution of co-occurrence
values, $N_{AB}$, changes more rapidly which changes in $\theta$ for taxon
pairs with greater numbers of occurrences, consistent with Figure S1. Thus,
$\theta$ can be estimated more efficiently from taxon pairs with more
occurrences. Importantly, however, Fisher Information eventually declines as the
numbers of occurrences for the two taxa approach the total number of sites (note
the depressed yellow curve for $N_{A} = N_{B} = 40$ in Figure S3). This decline
occurs due to an increasing lower bound on the minimum number of co-occurrences
($=\max(0,N_{B}-N_{(A))}$). Also noteworthy, the Fisher Information eventually
approaches 0 for very low and very high values of $\theta$, regardless of
$N_{A}$ and $N_{B}$, because the distributions of $N_{AB}$ values eventually
become concentrated on this lower bound as $\theta$ declines, and on the upper
bound $\min(N_{A},N_{B})$ as $\theta$ increases.


_Specifying the model_
The analysis above highlights three key issues that arise when fitting the NHD
to occurrence data:

(i)  A non-overlapping distribution (i.e.\ $N_{AB} = 0$) will be observed with
higher probability for a taxon pair with lower occurrence values ($N_{A}$ and
$N_{B}$), irrespective of $\theta$.

(ii)  A \emph{single} observation of a non-overlapping distribution will not 
\emph{by itself} be biologically informative about the value of $\theta$ (i.e.$\hat{\theta}=-\infty$).

(iii) The information about $\theta$ contained in a \emph{single} observation is
generally lower for a taxon pair with lower occurrence values.

In other words, the parameter $\theta$ suffers from similar constraints as existing metrics of species association. These constraints are seldom considered explicity by published co-occurrence analyses, except by throwing out pairs that do not yield biological information by themselves. To address these issues, we present a novel approach, which we term "occupancy analysis". Rather than separately estimate $\theta$ for each taxon pair, we instead \emph{combine} observations into occupancy groups,
$\mathcal{S}_i$, defined as having identical values for $N_{A}$ and $N_{B}$,
with $N_{A} \leq N_{B}$. We then calculate the likelihood for each occupancy
group based on a single $\theta$ value, $\theta_i$. This likelihood is simply
the product of the likelihoods for the co-occurrence observations in the
occupancy group, $N_{AB}^j$, with $j \in \mathcal{S}_i$. The likelihood
for a set of observations, $\vect{N_{AB}}$, is therefore

\begin{equation}
\tag{S4}
p(\vect{N_{AB}} \mid \theta_i) = \prod_{j \in \mathcal{S}_i} p(N_{AB}^j|\theta_i)
\end{equation}

This approach partly addresses issues (*i*) and (*ii*) because $N_{AB} = 0$ is less
likely to be observed for an entire set of observations than a single
observation, and the maximum likelihood estimate for $\theta_i$ will be finite
provided that at least 1 observation of $N_{AB}$ in the occupancy pattern is
$>\max(0,N_{B}-N_{(A)})$ and $<\min(N_{A},N_{B})$. It also addresses issue
(*iii*) because the Fisher Information is additive [@Ly2017],
meaning that it increases linearly with the number of observations.


To further address issues (*i*) and (*ii*), rather than separately calculating
$\theta_i$ for each set, we instead adopt a hierarchical Bayesian modeling
approach (Gelman, Carlin, Stern, & Rubin; Bayesian Data Analysis) to estimate
averages for $\theta_i$ that have been subdivided into four groups,
$g$, on the basis of habitat type (altered, unaltered) and diet (similar,
distinct). At the lower (i.e.\ set) level, we assume that the $\theta_i$ values
are all normally distributed with means and standard deviations that can vary
among groups

\begin{equation}
\tag{S5}
\theta_i \sim \mathcal{N}(\overline{\theta}_{g(i)}, \sigma_{g(i)})
\end{equation}

where $g(i)$ is a function that indexes the group assignment of each set. At the
upper (i.e.\ community) level, we assign the group means, $\overline{\theta}_g$,
a normal prior distribution with a mean of 0 and a standard deviation of 10,

\begin{equation}
\tag{S6}
\overline{\theta}_g \sim \mathcal{N}(0, 10)
\end{equation}

and we assign the group standard deviations, $\sigma_{g}$, a half-normal prior
distribution with location and scale parameters of 0 and 10, respectively, on
the interval $[0,\infty]$,

\begin{equation}
\tag{S7}
\sigma_{g} \sim \mathcal{H}(0, 10)
\end{equation}

These prior distributions were chosen to allow adequate exploration of the
parameter space using the Markov chain Monte Carlo (MCMC) algorithm while
ensuring that the algorithm never halted by getting stuck in regions of
parameter space with negligible likelihood. These prior distributions should be
viewed as very weakly informative in the parlance of Bayesian analysis [@Lemoine2019].


The posterior distributions for the group-level averages, $\overline{\theta}_g$,
are proportional to the product of two likelihood distributions (Eqs.\ S4 and
S5) and two prior distributions (Eqs.\ S6 and S7). Consequently, the model pools
data across sets of pairs to estimate the group means,
$\overline{\theta_g}$, and standard deviations, $\sigma_g$, and then, for each
set of occurrences, $i$, draws from a normal distribution of mean
$\overline{\theta_{g(i)}}$ and standard deviation $\sigma_{g(i)}$ to obtain a
$\theta_i$ estimate. Data pooling, in turn, results in shrinkage of estimates of
$\theta_i$ toward the population-level means, $\overline{\theta_g}$. A
hierarchical approach is desirable here because sets of pairs yielding weaker
evidence (as indexed by smaller likelihoods) exhibit greater shrinkage, and sets
yielding stronger evidence are assigned greater weight in calculating the
group-level means and standard deviations.

Occupancy analysis confers two major advantages over previous approaches. Though its estimates for single pairs suffer from the same constraints as classic co-occurrence analysis, it uses more information than co-occurrence analysis for groups of pairs, making its estimates of $\theta$ much more robust than, e.g., taking the mean co-occurrence value of a set of species pairs. Second, the algorithm is formulated in such a way that adding species does not increase the computing time of the model, meaning hyper-diverse datasets for which co-occurrence analysis is not feasible can be analysed with this method. 

_Fitting the model_
We estimate the posterior distributions using the R package RStan, which
provides an interface from R to the Stan probabilistic programming language (new
reference). To approximate the posterior distributions, we generate 4 MCMC
chains of $2 \times 10^3$ steps, including a warm-up period of $10^3$ steps. We
then inspect MCMC plots for the 4 chains and calculated Gelman Rubin statistics,
$\hat{R}$, to ensure convergence [@Gelman2013].


_Model robustness checks_
We ran a series of checks to test that the model is robust and not subject to common biases. First, we estimated the shrinkage present in our model results by comparing the bayesian model output to the same model calculated using maximum likelihood. Shrinkage refers to the tendency of observations with less information to gravitate toward the group mean in hierachical models. We want to see relatively little shrinkage in most sets, but more in occupancy sets with fewer pairs. We also want to see increased shrinkage in sets that fall farther from the group mean. These properties were observed for both bats and birds (Figs. S4-S5). 

We also examined our model output for near normal distribution and common biases. We plotted $\theta_g}$ estimates of each occupancy pair set against the number of pairs in the set and against the occupancy of the rarer species in the set. There were no consistent trends across the four experimental groups for either taxa, indicating that $\theta_g}$ does not suffer from bias due to these factors (Figs. S6-S11). 


_Model interpretation_

In this framework, a change in the posterior distributions of $\psi$ and therefore $\overline{\theta_g}$ can exhibit several patterns, which point to various biological processes (Fig. S12). Competing pairs can display lower $\overline{\theta_g}$ than non-competing pairs, which means that diffuse competition tends to lead to exclusion with respect to opportunity for co-occurrence (Fig. S12A). This is the case even if all $\overline{\theta_g}$ are positive, as exclusion does not necessarily present as a negative association. Because competing and non-competing pairs are comprised of the same species data in different pairwise arrangements, the $\overline{\theta_g}$ of non-competing pairs represents the opportunity for coexistence in the absence of food competition. A deviation toward lower $\overline{\theta_g}$ from this baseline is evidence that competing pairs are not coexisting as often as expected--a systematic pattern of exclusion at the group level. Conversely, a higher $\overline{\theta_g}$ in competing pairs confirms that similarity in resource use leads to co-occurrence without causing exclusion, suggesting that competing pairs tend to partition or share adequate resources (Fig. S12B). If the $\overline{\theta_g}$ of competing and non-competing pairs do not differ significantly, but the $\sigma_{g}$ (variance) of competing pairs is higher, this points to stronger spatial patterns (both coexistence and exclusion) in pairs that compete for food, and vice versa (Fig. S12D). If there is no difference in $\overline{\theta_g}$ or $\sigma_{g}$ between the two groups, then there is insufficient evidence to support the hypothesis that diffuse competition is influencing the structure of the assemblage at the landscape scale (Fig. S12C).

The $\overline{\theta_g}$ and $\sigma_{g}$ of competing and non-competing pairs might also differ between altered and intact sites in our example. If the relationship between competing and non-competing pairs is different, this means that alteration has changed the outcome of biotic interactions, e.g., by promoting partitioning over exclusion or vice versa. More subtly, the extent of disparity between the two groups could differ without qualitatively changing the relationship, which indicates that naturally existing mechanisms have been enhanced or dampened. On the other hand, $\overline{\theta_g}$ or $\sigma_{g}$ could simply be higher or lower across the board, suggesting that alteration leads to changes in community structure that do not necessarily involve the rewiring of competitive interations. In all cases, posteriors represent assemblage-level patterns and do not imply that all pairs in the group exhibit the same pattern as does their group, merely that there is a group tendency toward one process over another.

The experimental approach presented here represents two distinct innovations that allow us to deal with detection issues and context-dependence of interactions. First, we use a priori evidence for the possibility of interaction rather than direct interaction data. The resulting set of pairs includes both interacting and non-interacting pairs but encompasses all of the former. A priori evidence for many types of interactions could be derived from various functional traits. For example, fruit size and gape size of frugivorous birds can indicate which birds are physically able to disperse which fruits. Diet category and body size can similarly be used to estimate mammal predator-prey relationships [@Pires2015]. This setup frees us from the necessity of sampling all interactions and identifying the contexts under which they are realised while still yielding insights about the impact of interactions on community assembly.

Second, we use non-interacting pairs of species in the assemblage to contextualise association patterns for interacting species pairs. The inclusion of an experimental control group is a central tenet of science, but non-interacting pairs or those having associations not significantly different from chance associations are routinely discarded or ignored in co-occurrence research and interaction research more generally, while raw pairwise co-occurrence scores are often interpreted at face value (e.g., any negative score equals exclusion). Our use of non-interacting pairs as a control can be applied to a broad variety of interactions. A plant-pollinator network could for instance be decomposed into groups of pairs that facilitate one another (plant-pollinator pairs), do not interact, and even compete (plant-plant pairs that compete for pollinators or pollinator-pollinator pairs that compete for nectar). In each case, the effect of the competing group(s) can be evaluated against all pairs that do not exhibit the interaction(s) in question.

_The effect of turnover_

Overall changes observed in association patterns could result from pairs changing their spatial behaviour or their competition intensity in altered vs. intact habitats. Conversely, a shift in species composition (e.g., the removal or introduction of interaction partners), and therefore the presence of a different suite of pairs, could also cause disparities between the two habitat types. To investigate these possibilities, we ran the models with all species and then repeated them with only species occurring at least once in both habitat types ("no-turnover" models). If the results of the two model sets are very similar, we can conclude that the results are driven by changes in the spatial behaviour of species occurring in both site types. If the two model sets are different but both show significant effects, likely both turnover and changes in preserved interactions are occurring. Furthermore, the differences between the two outputs can be used to determine which changes are due to which mechanism. Model effects are due to turnover alone if the model set using only shared species has no significant differences while the model set using all species has significant differences.

_Analysis of alteration types_ 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# calculate number of species per guild (in the functional analysis)
s <- spp %>% filter(rownames(.) %in% unlist(map(PAn, rownames))) %>% group_by(life.form, guild) %>% dplyr::summarise(count = length(guild))

```

To gain more insight into the aggregation of same-guild (competing) pairs in altered habitats, we calculated the percent of species in each guild that co-occur at individual sites (e.g., 5% of insectivorous bat species occur at site A, etc.). We plotted the sites by alteration type to examine the influence of alteration on species aggregation. Although we do not have enough sites of each alteration type to draw any definitive conclusions, this analysis can suggest reasons why certain guilds might co-occur more in altered sites and provide questions for future research.

### Supplementary Text
```{r echo= FALSE, message=FALSE, warning=FALSE}
# occupancy calculations
occ <- map(PAn, t)  %>% map(data.frame) %>% map(~split(., rownames(.) %in% unalt_sites)) %>% 
  map(map, ~colSums(.)/nrow(.)) %>% map(map, data.frame) %>% map(~merge(.[[1]], .[[2]], by = 0)) %>% map(setNames, c("name", "altered", "intact")) %>% bind_rows(.id = "taxon")
docc <- occ %>% group_by(taxon) %>% summarise(
  declined = length(which(altered < intact)), 
  increased = length(which(intact < altered)), 
  extirpated = length(which(altered == 0 & intact > 0)), 
  appeared = length(which(altered > 0 & intact == 0)), 
  shared = length(which(altered > 0 & intact > 0)), 
  total = length(altered))
```

_Richness_ 

Richness was calculated on the biogeographically matched sites, so the number of altered and intact sites is comparable. Based on Chao 1 and cJ1, altered and intact sites have similar alpha richness (cJ1 _p_-values = 
`r rsig %>% filter(metric == "cJ1", taxon == "bat") %>% pull(p) %>% round(3)` and 
`r rsig %>% filter(metric == "cJ1", taxon == "bird") %>% pull(p) %>% round(3)`, and Chao 1 _p_-values = 
`r rsig %>% filter(metric == "chao", taxon == "bat") %>% pull(p) %>% round(3)` and 
`r rsig %>% filter(metric == "chao", taxon == "bird") %>% pull(p) %>% round(3)` for bats and birds, respectively). 
However, Fisher's alpha indicated a significant diversity decrease in bats at altered sites ( _p_-value = 
`r rsig %>% filter(metric == "fa", taxon == "bat") %>% pull(p) %>% round(3)`).
Average richness for bat samples in intact habitats was 
`r rsumm %>% filter(metric == "cJ1", taxon == "bat", status == "Intact") %>% pull(mean.rich) %>% round(2)` (cJ1) and 
`r rsumm %>% filter(metric == "chao", taxon == "bat", status == "Intact") %>% pull(mean.rich) %>% round(2)` (Chao1), and
`r rsumm %>% filter(metric == "cJ1", taxon == "bat", status == "Altered") %>% pull(mean.rich) %>% round(2)` (cJ1) and 
`r rsumm %>% filter(metric == "chao", taxon == "bat", status == "Altered") %>% pull(mean.rich) %>% round(2)` (Chao 1) in altered habitats. Average richness for bird samples  was 
`r rsumm %>% filter(metric == "cJ1", taxon == "bird", status == "Intact") %>% pull(mean.rich) %>% round(2)` (cJ1) and 
`r rsumm %>% filter(metric == "chao", taxon == "bird", status == "Intact") %>% pull(mean.rich) %>% round(2)` (Chao1), and
`r rsumm %>% filter(metric == "cJ1", taxon == "bird", status == "Altered") %>% pull(mean.rich) %>% round(2)` (cJ1) and 
`r rsumm %>% filter(metric == "chao", taxon == "bird", status == "Altered") %>% pull(mean.rich) %>% round(2)` (Chao 1) in altered habitats.

For bats and birds, respectively, 
`r round(docc[1,3] / docc[1,7]*100)`% and 
`r round(docc[2,3] / docc[2,7]*100)`% of species increased their occupancies in altered habitats, and of these, 
`r round(docc[1,5] / docc[1,3]*100)`% and 
`r round(docc[2,5] / docc[2,3]*100)`%, respectively, appeared only in altered habitats. Overall, however, 
`r round(docc[1,2] / docc[1,7]*100)`% of all bat and 
`r round(docc[2,2] / docc[2,7]*100)`% of all bird species decreased their occupancies in altered habitats, and of these, 
`r round(docc[1,4] / docc[1,2]*100)`% and 
`r round(docc[2,4] / docc[2,2]*100)`% disappeared altogether. A surprising number of species were sampled only in altered habitats (`r round(docc[1,5]/ docc[1,7]*100)`% of bats and `r round(docc[2,5]/ docc[2,7]*100)`% of birds). These species may exist in intact habitats. However, as we took measures to standardise sampling between altered and intact sites, these species likely represent a group whose abundances have been substantially increased in altered sites. The converse may be true for some species that were only sampled in intact sites (`r round(docc[1,4]/ docc[1,7]*100)`% of bats and `r round(docc[2,4]/ docc[2,7]*100)`% of birds).

_Beta diversity and composition_
```{r message=FALSE, warning=FALSE, include=FALSE}
pbat <- beta[[1]]$`Pr(>F)`[1]
pbird <- beta[[2]]$`Pr(>F)`[1]
```

There was no significant disparity in the beta diversity of altered and intact sites for either taxon (Table S1), regardless of the metric used (ANOVA of multivariate dispersion, _p_~bat~ = 
`r output %>% filter(analysis == "beta dispersion", metric == "Ochiai") %>% pull(p.bat)` and _p_~bird~ = 
`r output %>% filter(analysis == "beta dispersion", metric == "Ochiai") %>% pull(p.bird)` for the Ochiai index;  _p_~bat~ = 
`r output %>% filter(analysis == "beta dispersion", metric == "Cosine similarity") %>% pull(p.bat)` and _p_~bird~ = 
`r output %>% filter(analysis == "beta dispersion", metric == "Cosine similarity") %>% pull(p.bird)` for the cosine similarity). 
Both multivariate permutation (regardless of the metric) and canonical correspondence analysis detect a shift in the composition of bird communities between altered and intact habitats, and both yield stronger _p_-values when abundance data are included (Table S1). Results of composition analyses are inconsistent for bat communities, suggesting significant compositional change only when abundance data are used in canonical correspondence analysis (Table S1). 

_Interaction results_ 

Occupancy model results are covered extensively in the main text. See Table S1 for 95% high-density posterior intervals for all model runs. 

We re-ran our models excluding species that were not sampled in both altered and intact habitats (no-turnover models). The purpose was to discover whether the main drivers of the observed patterns are (a) compositional disparities between the two habitat types, (b) actual changes in the spatial behavior of common species between the habitat types, or (c) both. There were strong similarities between these no-turnover models and our original models. In bats, there was very little difference between the two model sets (compare Fig. 3A to 3B), supporting hypothesis (b). In birds, the same relationships are observed between habitats, but competing pairs in altered habitats have a $\bar{\theta{g}}$ equal to control pairs (Fig. 4A-B) compared to the original model sets (Fig. 4A), in which competing pairs have a slightly lower median $\bar{\theta{g}}$. This suggests that competing species occurring only in altered habitats co-occur less than those occurring in both habitats and supports hypothesis (c). However, competing pairs still exhibit coexistence in intact habitats, and the relationship is still lost in altered habitats. Altered habitats cannot consistently sustain diverse sets of competing pairs, irrespective of whether the species comprising them occur in both types of habitats or primarily in altered habitats.


_Altered site types_

 Strong aggregations form when species occur together at several sites and are both absent from other sites. To help understand our results, we calculated the representation of dietary guilds at various different types of altered sites (Fig. S13). Carnivore-insectivore (9 species) and frugivore bats (69 species) co-occur disproportionately in plantations. Plantations and forest fragments support the highest proportion of insectivore bats (98 species) living in altered habitats, and nectarivores (24 species) co-occur disproportionately in plantations and croplands (Fig. S13).
 
For birds, the single site in a zone of human habitation supports a disproportionate ratio of birds with simple diets (nectarivores, frugivores, invertivores, and granivores). Croplands are the main supporter of frugivore-granivore birds. Many of our alteration categories included only one or two sites, which is insufficient to draw definitive conclusions about how bat and bird guilds aggregate over the altered landscape. However, our results (Fig. S13) suggest that plantations are an important aggregation point for bats, particularly frugivores and nectarivores. The data for site types are even sparser for birds, but the primary habitat types are forest fragment, combined, disturbed forest, and plantation. These site types vary in their importance to individual bird guilds. Birds with fruit or nectar in their diets showed a variety of site preferences, while insectivores were most strongly aggregated in combined and inhabited sites.


```{r, echo = FALSE, include = TRUE, fig.height=5.5, fig.width = 7}
#load("../Data/coords.Rdata")

### FIGURE S1: Maps ####
par(mfrow = c(2,2), mar = c(0,0,0,0), oma = c(1,1,1,1))

maps::map("world", xlim = c(-125, -30), ylim = c(-40, 25))
points(latitude~longitude, coords1$bat, pch = 16, cex=.7, col = "#006D77")
points(latitude~longitude, coords2$bat, pch = 16, cex=.7, col = "#DF8768")
mtext("Bats-original sites", side = 3, line = 1)
box()

maps::map("world", xlim = c(-125, -30), ylim = c(-40, 25))
points(latitude~longitude, coords1$bird, pch = 16, cex=.7, col = "#006D77")
points(latitude~longitude, coords2$bird, pch = 16, cex=.7, col = "#DF8768")
mtext("Birds-original sites", side = 3, line = 1)
box()

maps::map("world", xlim = c(-125, -30), ylim = c(-40, 25))
points(latitude~longitude, coords1.keep$bat, pch = 16, cex=.7, col = "#006D77")
points(latitude~longitude, coords2.keep$bat, pch = 16, cex=.7, col = "#DF8768")
mtext("Bats-included sites", side = 3, line = 1)
box()

maps::map("world", xlim = c(-125, -30), ylim = c(-40, 25))
points(latitude~longitude, coords1.keep$bird, pch = 16, cex=.7, col = "#006D77")
points(latitude~longitude, coords2.keep$bird, pch = 16, cex=.7, col = "#DF8768")
mtext("Birds-included sites", side = 3, line = 1)
box()


```

__Fig. S1.__ 
Maps of Neotropical bat and bird sites before (top) and after (bottom) biogeographic matching procedure, with altered sites represented in red and intact sites represented in blue.

```{r figs2, dpi=200, echo=FALSE, fig.cap="\\label{fig:figs2}Effects of taxon occurrence (indexed by $N_{A}$ and $N_{B}$) and spatial aggregation ($\\theta>0$) and segregation ($\\theta<0$) on the probability distribution of co-occurrences, $f(N_{AB} \\mid \\theta)$. Probabilities were calculated using Eq.\\ (S1) based on a total of $N_{A} + N_{(A)}=50$ sites. Probabilities $<10^{-3}$ are not plotted."}
library(scales)

log_sum_exp <- function(x) log(sum(exp(x - max(x)))) + max(x)
fnchypergeo_lpmf <- function(sb, s1, s2, n_site, theta) {
  sb_min <- max(c(0,s1 + s2 - n_site))
  sb_max <- min(c(s1,s2))
  u <- ll <- ans <- numeric()
  for(i in sb_min:sb_max) {
    u[i - sb_min + 1] = i;
    ll[i - sb_min + 1] = lchoose(s1, i) + lchoose(n_site - s1, s2 - i)
  }
  if(length(sb) > 1) {
    for(i in 1:length(sb)) {
      ans[i] = ll[sb[i] - sb_min + 1] + sb[i]*theta - log_sum_exp(ll + u*theta)
    }
  } else {
    for(i in 1:length(theta)) {
      ans[i] = ll[sb - sb_min + 1] + sb*theta[i] - log_sum_exp(ll + u*theta[i])
    }
  }    
  return(ans)
}


ll_plot <- function(s1, s2, n_site, theta,...) {
  sb_min <- max(c(0,s1 + s2 - n_site))
  sb_max <- min(c(s1,s2))
  x <- sb_min:sb_max
  y <- fnchypergeo_lpmf(sb_min:sb_max, s1, s2, n_site, theta)
  lines(x,y,lwd=1.8,type='o',pch=16,cex=0.5,...)
}

my.cols <- c('red','blue')
plot(0,0,xlim=c(0,9),ylim=c(-7,0),type='n',
     xlab=expression("Number of co-occurrences,"~italic(N[AB])),
     ylab = expression("Probability,"~italic(f)(italic(N[AB])~"|"~italic("\u03B8"))),
     axes=FALSE)
axis(1,at=0:9,labels=0:9)
axis(2,at=log(10^seq(-3,0,1)),labels=parse(text=paste0('10^{',seq(-3,0,1),'}')),
     las=1)
box()
ll_plot(2,2,50,-2,col=alpha(my.cols[1],0.5),lty=3) 
ll_plot(2,2,50,0,col=alpha(my.cols[1],0.5),lty=2)
ll_plot(2,2,50,2,col=alpha(my.cols[1],0.5),lty=1)
ll_plot(8,8,50,-2,col=alpha(my.cols[2],0.5),lty=3) 
ll_plot(8,8,50,0,col=alpha(my.cols[2],0.5),lty=2)
ll_plot(8,8,50,2,col=alpha(my.cols[2],0.5),lty=1)
legend('topright',bty='n',lwd=1.8,
       legend=expression(italic("\u03B8")~"="~-2,
                         italic("\u03B8")~"="~0,
                         italic("\u03B8")~"="~2,'',
                         italic(N[A])~"="~italic(N[B])~"="~2,
                         italic(N[A])~"="~italic(N[B])~"="~8),
       lty=c(3,2,1,0,1,1),
       col=c(rep('black',4),my.cols))
```


```{r figs3, dpi=200, echo=FALSE, fig.cap="\\label{fig:figs3}Effects of occurrence ($N_{A}$ and $N_{B}$) and spatial aggregation ($\\theta$) on the probability distribution of co-occurrences. Probabilities were calculated using Eq.\ (S3), assuming a total of $N_{A} + N_{(A)}=50$ sites."}
log_sum_exp <- function(x) log(sum(exp(x - max(x)))) + max(x)
fnchypergeo_FI <- function(s1, s2, n_site, theta) {
  sb_min <- max(c(0,s1 + s2 - n_site))
  sb_max <- min(c(s1,s2))
  ans<- numeric()
  u <- ll <- ans <- numeric()
  for(i in sb_min:sb_max) {
    u[i - sb_min + 1] = i;
    ll[i - sb_min + 1] = lchoose(s1, i) + lchoose(n_site - s1, s2 - i)
  }
  for(i in 1:length(theta)) {
    ans[i] <-
      exp(log_sum_exp(2*log(u) + ll + u*theta[i])-log_sum_exp(ll + u*theta[i])) - 
      exp(log_sum_exp(log(u) + ll + u*theta[i])-log_sum_exp(ll + u*theta[i]))^2
  }
  return(ans)
}

o.sq <- seq(-10,10,.01)
plot(o.sq,fnchypergeo_FI(2, 2, 50, o.sq),las=1,
     xlab=expression("NHD Parameter, "~italic("\u03B8")),type='l',ylim=c(0,3),
     ylab=expression("Fisher Information, "~italic(I)(italic("\u03B8"))),
     col=hcl.colors(5)[1],lwd=1.5)
lines(o.sq,fnchypergeo_FI(2, 8, 50, o.sq),
      lty=2,col=hcl.colors(5)[1],lwd=1.5)
lines(o.sq,fnchypergeo_FI(8, 8, 50, o.sq),
      lty=3,col=hcl.colors(5)[2],lwd=1.5)
lines(o.sq,fnchypergeo_FI(8, 16, 50, o.sq),
      lty=4,col=hcl.colors(5)[2],lwd=1.5)
lines(o.sq,fnchypergeo_FI(16, 16, 50, o.sq),
      lty=5,col=hcl.colors(5)[3],lwd=1.5)
lines(o.sq,fnchypergeo_FI(32, 32, 50, o.sq),
      lty=6,col=hcl.colors(5)[4],lwd=1.5)
lines(o.sq,fnchypergeo_FI(40, 40, 50, o.sq),
      lty=7,col=hcl.colors(5)[5],lwd=1.5)
legend('topleft',legend=expression(italic(N[A])~"="~italic(N[B])~"="~2,
                                   italic(N[A])~"="~2~", "~italic(N[B])~"="~8,
                                   italic(N[A])~"="~italic(N[B])~"="~8,
                                   italic(N[A])~"="~8~", "~italic(N[B])~"="~16,
                                   italic(N[A])~"="~italic(N[B])~"="~16,
                                   italic(N[A])~"="~italic(N[B])~"="~32,
                                   italic(N[A])~"="~italic(N[B])~"="~40),
       lty=1:7,bty='n',col=hcl.colors(5)[c(1,1,2,2,3,4,5)],lwd=1.5,cex=0.75)
```

```{r figs4, echo = FALSE, include = TRUE, fig.height=7, fig.width = 5, dpi = 200}
data_bat_full <- readRDS("../Results/stan/Model_data_bats_all_full.rds")
bat_full <- readRDS("../Results/stan/stan6_bats_all_full.rds")
ml_bat_full <- readRDS("../Results/stan/ML_bats_all_full.rds")
sum_bat <- stan_summary(data_bat_full, bat_full, ml_bat_full)
stan_shrinkage_plots(sum_bat)

```

__Fig. S4.__ Results of shrinkage tests for bat model with all species, depicting shrinkage with respect to deviation from the group mean and with respect to the number of pairs in each set.

```{r figs5, echo = FALSE, include = TRUE, fig.height=7, fig.width = 5, dpi = 200}
data_bird_full <- readRDS("../Results/stan/Model_data_birds_all_full.rds")
bird_full <- readRDS("../Results/stan/stan6_birds_all_full.rds")
ml_bird_full <- readRDS("../Results/stan/ML_birds_all_full.rds")
sum_bird <- stan_summary(data_bird_full, bird_full, ml_bird_full)
stan_shrinkage_plots(sum_bird)
```

__Fig. S5.__ Results of shrinkage tests for bird model with all species, depicting shrinkage with respect to deviation from the group mean and with respect to the number of pairs in each set.

```{r figs6, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bat, aes(x =theta_bayes)) + 
  geom_histogram(bins = 12) + 
  facet_wrap(~group) + labs(x = "theta")

```

__Fig. S6.__ Distribution of theta posteriors for each group in the bat model. 

```{r figs7, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bat, aes(x = s1, y = theta_bayes)) + 
  geom_point(size = .5, alpha = .5) + facet_wrap(~group) + 
  geom_smooth(method = "loess") + 
  geom_line(aes(y = 0), lty = 2) + 
  labs(x = "occupancy of rarer species in pattern")

```

__Fig. S7.__ Theta estimates for bats plotted against the occupancy of the rarer species in the occupancy pattern. Each point represents a pair set with a particular occupancy pattern. Each group is plotted separately.

```{r figs8, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bat, aes(x = log_n, y = theta_bayes)) + 
  geom_point(size = .5, alpha = .5) + facet_wrap(~group) + 
  geom_smooth(method = "loess") + 
  geom_line(aes(y = 0), lty = 2) + 
  labs(x = "log(number of pairs)")

```

__Fig. S8.__ Theta estimates for bats plotted against the number of pairs in the set, with each point representing one pair set with a particular occupancy pattern. 

```{r figs9, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bird, aes(x =theta_bayes)) + 
  geom_histogram(bins = 12) + 
  facet_wrap(~group) + labs(x = "theta")

```

__Fig. S9.__ Distribution of theta posteriors for each group in the bird model. 

```{r figs10, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bird, aes(x = s1, y = theta_bayes)) + 
  geom_point(size = .5, alpha = .5) + facet_wrap(~group) + 
  geom_smooth(method = "loess") + 
  geom_line(aes(y = 0), lty = 2) + 
  labs(x = "occupancy of rarer species in pattern")

```

__Fig. S10.__ Theta estimates for birds plotted against the occupancy of the rarer species in the occupancy pattern. Each point represents a pair set with a particular occupancy pattern. Each group is plotted separately.

```{r figs11, echo = FALSE, include = TRUE, fig.height=5, fig.width = 5, dpi = 200}
ggplot(sum_bird, aes(x = log_n, y = theta_bayes)) + 
  geom_point(size = .5, alpha = .5) + facet_wrap(~group) + 
  geom_smooth(method = "loess") + 
  geom_line(aes(y = 0), lty = 2) + 
  labs(x = "log(number of pairs)")

```

__Fig. S11.__ Theta estimates for birds plotted against the number of pairs in the set, with each point representing one pair set with a particular occupancy pattern. 

__Fig. S12.__ Interpretations of average $\overline{\theta_g}$ posterior density distributions. The dark curve indicates competing pairs and the lighter curve indicates control pairs. In (A) competing pairs co-occur less on average than control pairs, so higher spatial segregation than expected suggesting a pattern of competitive exclusion. In (B) competing pairs co-occur more than control pairs on average, suggesting that they micro-partition resources at the same sites more than expected. In (C) there is little or no separation of the posterior distributions, meaning that food competition does not aggregate or segregate pairs outside of expectations; however, other interaction types could still play an important role. (D) Both strong aggregation and segregation are observed in competing pairs, so variance is higher than expected.

```{r, echo = FALSE}
#load("../Results/d2_d5_summary_tables_nr_ns_100r_relatedexcl.RData")
colors2 <- c("#34008C", "#E8AC00")
specs <- list(geom_hline(yintercept = 1, col = "gray"),
              geom_vline(xintercept = 1, col = "gray"),
              stat_ellipse(geom= "polygon", alpha = 0.3, aes(col = diet.match, fill = diet.match)),
              geom_point(aes(col = diet.match, fill = diet.match), size = 0.5), 
              geom_rug(aes(col = diet.match), alpha = 0.5),
              scale_colour_manual(values = colors2),
              scale_fill_manual(values = colors2),
              theme(legend.position = "none")
)

```


```{r, echo = FALSE, message=FALSE, warning=FALSE, include = TRUE, fig.height = 8, fig.width = 11}
load("../Results/pr_bats_birds_percent_spp_supported_by_sites.RData")

pbat <- ggplot(pr.bat, aes(x = altered_habitat, y = value*100, fill= altered_habitat)) + geom_boxplot(outlier.size = 0.3) +
  facet_wrap(variable~., nrow = 4, ncol = 1, scales = "free_y") + geom_jitter(width = 0.15, size = 0.3) +
  scale_fill_manual(values = c(gg_color_hue(8), "gray50")) +
  theme(axis.text.x = element_blank()) + labs(y = "Percent of species co-occurring", x = "Habitat type")


pbird <- ggplot(pr.bird, aes(x = altered_habitat, y = value*100, fill= altered_habitat)) + geom_boxplot(outlier.size = 0.3) +
  facet_wrap(variable~., nrow = 4, ncol = 2, scales = "free_y") + geom_jitter(width = 0.15, size = 0.3) +
  theme(axis.text.x = element_blank()) + labs(y = "Percent of species co-occurring", x = "Habitat type", fill = "Type of alteration") + 
  scale_fill_manual(values = c(gg_color_hue(8), "gray50"))

plot_grid(pbird+ theme(legend.position = c(1, 0), legend.justification = c(1, 0), legend.text = element_text(size = 10)) + guides(fill=guide_legend(ncol=2)), 
          pbat+theme(legend.position="none"), ncol = 2, rel_widths = c(2, 1), labels = c("A", "B"))

```

__Fig. S13.__ 
Percent of species in (A) bird and (B) bat guilds, represented by panels, that co-occur at individual altered sites by type of alteration. Each point is a site. Key to panels: C = carnivore; F = frugivore; I = insectivore; N = nectarivore; and G = granivore. Multiple letters indicate mixed feeders.

__Table S1.__
Results of beta diversity and composition analyses comparing community patterns in altered and intact habitats. F-statistic and _p_-values are displayed for bats and birds.

```{r echo=FALSE, message=FALSE, warning=FALSE}
output %>% arrange(analysis) %>% mutate(across(where(is.numeric), round, digits = 4)) %>% flextable()
```


__Table S2.__ 
Median values and 95% high density posterior intervals for $\overline{\theta_{g}}$ and $\sigma_{g}$ of all bat and bird models.
```{r echo=FALSE, message=FALSE, warning=FALSE}

l <- list.files('../Results/stan', full.names = TRUE, pattern = "stan6")

res <- map(l, ~readRDS(.x) %>% format_stanfit("mu")) %>%
  setNames(c(paste(ifelse(grepl("bat",l, fixed = TRUE), "bat", "bird"),
                   ifelse(grepl("shared", l, fixed = TRUE), "noTurnover", "full")) %>% trimws() %>% str_replace_all(" ", "_"))) %>% 
  bind_rows(.id = "model")

res2 <- map(l, ~readRDS(.x) %>% format_stanfit("sigma")) %>%
  setNames(c(paste(ifelse(grepl("bat",l, fixed = TRUE), "bat", "bird"),
                   ifelse(grepl("shared", l, fixed = TRUE), "noTurnover", "full")) %>% trimws() %>% str_replace_all(" ", "_"))) %>% 
  bind_rows(.id = "model")

res$sigma <- res2$sigma

hdpi95 <- res %>% group_by(model, group) %>% 
  summarise(`Median θ` = median(mu), `Lower θ` = hdi(mu, 0.95)$CI_low, `Upper θ` = hdi(mu, 0.95)$CI_high, `Median σ` = median(sigma), `Lower σ` = hdi(sigma, 0.95)$CI_low, `Upper σ` = hdi(sigma, 0.95)$CI_high) %>% mutate(across(where(is.numeric), round, digits = 4)) %>%
  separate(group, into = c("status", "Pair_type")) %>%
  separate(model, into = c("taxon", "model"), extra = "merge") 
  
flextable(hdpi95)

```

__References__

