---
title: The extirpation of biological interactions from human-altered habitats
output:
  bookdown::word_document2: default
bibliography: References.bib
csl: science.csl

---

__Authors:__ Anikó B. Tóth^1^*, Andrew P. Allen^2^, S. Kathleen Lyons^3^, John Alroy^2^

__Affiliations:__

^1^ Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW, Sydney, NSW 2052, Australia.

^2^ Department of Biological Sciences, Macquarie University, New South Wales 2109, Australia.

^3^ School of Biological Sciences, University of Nebraska-Lincoln, Lincoln, Nebraska 68588, USA.

*Correspondence to aniko.toth@unsw.edu.au

__Abstract:__ Anthropogenic habitat destruction is among the greatest threats facing animal communities. Increasing evidence points to a disruption of biotic interactions that drive community assembly even at broad spatial scales, jeopardizing ecosystem function. Here, we assess whether spatial patterns of co-occurrence among species pairs that compete for food resources differ between altered and intact habitats for Neotropical bird and bat assemblages. We find that competing species pairs are less spatially aggregated in altered habitats. Our results suggest a loss of ecosystem functionality, specifically the ability of altered habitats to support diverse competitive interactions. Habitat alteration therefore strongly affects community assembly via changes in competitive outcomes, reversing patterns observed in the wild.

__One sentence summary:__ Coexistence of competing species of Neotropical bats and birds is reduced in human-altered habitats.

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(igraph)
library(reshape2)
library(vegan)
library(sp)
library(stringi)
library(parallel)

knitr::opts_chunk$set(include = FALSE)

## Load helper functions
source('../Code/HelperFunctions.R')
## Prep raw data
source('../Code/Data_Prep.R')

# species metadata
#load("../Data/Species_metadata.RData")
#load("../Data/cosmo_species_cosmopolitan_score.RData") # cosmopolitan category 
# Site data #
#load("../Data/sitedat_metadata.Rdata")

## Richness results
load("../Results/cj1_chao_richness_raw.RData")

taxa = c("bat", "bird")

unalt_sites <- sitedat[sitedat$altered.habitat == "",]$sample.no

```


__Main text:__
Human societies rely on ecosystem services, many of which depend in turn on a balance of interspecific interactions. For example, population control of a destructive pest species or disease vector may be contingent on competitive and/or predatory interactions with other species. Unprecedented global human impacts may be putting ecosystem services at risk, but empirical studies quantifying how biotic interactions change under ecosystem degradation are limited. The study of ecological networks of interactions is one active research area that may prove useful to understanding the effects of human impacts on species interactions [@Baiser2019]. Indeed, there is evidence that human disturbance can decouple predator-prey interactions [@Rodewald2011], increase competition for pollinators [@Holzschuh2011], and simplify or break apart interaction networks [@Haddad2000; @DeAssisBomfim2018].

Though still contentious [@McGill2010; @Araujo2013; @Ezard2016; @Marshall2016], it is increasingly recognized that biotic interactions can scale up to influence community assembly at much larger spatiotemporal scales [@Araujo2007; @Pollock2014; @Fraser2020] than that of individual encounters. For example, indirect evidence suggests that interactions such as competition and facilitation can influence the continental distribution of taxa over thousands of years [@Toth2019], and drive the diversification and extinction of entire clades over millions of years [@Sepkoski1984; @Alroy2010]. 

The details of how this upscaling unfolds is still not well understood because most detailed investigations of interspecific interactions are limited in terms of species and/or spatial extent due the challenges involved in studying them. Biotic interactions may be under-sampled or missed altogether due to inadequate field observations. Interactions are also context-dependent because ecological circumstances can influence whether an interaction takes place and its outcome [@Song2020]. For instance, species pairs can exhibit stable or unstable population dynamics  depending on the presence or absence of an apex predator in the system [@Karakoc2020], or switch from facilitation to competition along an environmental gradient [@Bimler2018]. Competitive interactions can result in coexistence or exclusion, depending on fitness, niche disparity, and priority effects of the species involved [@Grainger2019]. Human habitat alteration in particular has a vast potential to change ecological context and affect the dynamics of biotic interactions, all without necessarily adding or removing species [@Tylianakis2017].

```{r, include = FALSE}
#### Change to presence-absence ####
PAn<- tobinary(PAn)
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites])) # separate intact
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites])) # and altered sites

#### Match Biogeography ####
meta <- sitedat %>% split(.$taxon)
coords <- map(meta, select, c(sample.no, latitude, longitude)) 

# Remove intact or altered sites that are not near a site of the other type. This is done to ensure the two sets have similar biogoegraphical distributions.
coords1 <- purrr::map2(coords, PAnu, function(x, y) x[x$sample.no %in% colnames(y),])
coords2 <- purrr::map2(coords, PAna, function(x, y) x[x$sample.no %in% colnames(y),])
keep <- map2(coords1, coords2, matchbiogeo) %>% map(unlist)

PAn <- map2(PAn, keep, function(x, y) return(x[,as.character(y)]))
PAn <- map(PAn, clean.empty, minrow = 1) # remove any species that now have no occurrences

# recalculate alt/unalt split from new PAn
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites]))
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites]))

# coords of sites we kept, for plotting later.
coords1.keep <- purrr::map2(coords, PAnu, function(x, y) x[x$sample.no %in% colnames(y),])
coords2.keep <- purrr::map2(coords, PAna, function(x, y) x[x$sample.no %in% colnames(y),])

```

Herein, we examine how human alteration of habitat influences a specific type of interspecific interaction (food competition) through its effects on spatial distributions of species pairs (co-occurrence vs segregation). Our statistical modeling approach deals flexibly with context-dependence (e.g. whether or not an interaction is actualized in practice) by including all potential interactions and has the capacity to elucidate how associations between individual pairs of species collectively result in assemblage-level spatial patterns. We apply the model to assemblages of neotropical bats and birds downloaded from the Ecological Register [@Alroy2015] consisting of `r sum(PAn[["bat"]])` occurrences of bats and `r sum(PAn[["bird"]])` occurrences of birds at `r ncol(PAn[["bat"]])` and `r ncol(PAn[["bird"]])` sites, respectively. Our analysis demonstrates that habitat degradation influences spatial patterns of co-occurrence among competitors at regional to continental scales by altering–even reversing–the average outcome of food competition across regional assemblages. Traditional analyses of community structure are unable to capture these results, so changes are not reflected in richness, composition, and beta-diversity analyses.

```{r, echo = FALSE}
unalt_sites <- sitedat %>% filter(status == "Unaltered") %>% pull(sample.no) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}

# composition PCoA

dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "sample.no"))

# Significance test on centroids
w <- numeric()
for(j in 1:length(ord)){
  i <- ord[[j]]
  d <- numeric()
  for(r in 1:1000){
    i$rand = round(runif(n = nrow(i), min = 0, max = 1), 0)
    cent <- i %>% group_by(rand) %>% dplyr::summarise(x = mean(X1), y = mean(X2))
    d[r] <- spDists(as.matrix(cent[,c(2:3)]))[2]
  }
  cent <- i %>% group_by(status) %>% dplyr::summarise(x = mean(X1), y = mean(X2))
  obs <- spDists(as.matrix(cent[,c(2:3)]))[2]
  w[j] <- wilcox.test(d, obs)$p.value
}
w <- round(w, 3)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
### Beta - run using presence-absence, biogeo matched data ####

beta <- beta.types(PAn, unalt_sites)

# summary stats
b <- beta %>% group_by(taxon, unalt.pair) %>% summarise(mean = mean(Z.Score), median = median(Z.Score))
# p - values
bdiv <- beta %>% dplyr::filter(unalt.pair != "Unaltered-Altered") %>% 
  mutate(similarity = qnorm(Z.Score)) %>% split(.$taxon) %>% 
  purrr::map(~wilcox.test(data = ., similarity~unalt.pair, paired=FALSE))

```

There are no significant differences in the composition of altered and intact sites (Fig. 1) for bats ($p$= `r w[1]`) or birds ($p$= `r w[2]`). Intact and altered habitats do not differ significantly in species richness, but bat beta diversity is lower in altered habitats (Wilcoxon rank sum test, $p_{bird}$= `r round(bdiv$bird$p.value, 4)`, $W_{bird}$= `r round(bdiv$bird$statistic, 4)`, $p_{bat}$= `r round(bdiv$bat$p.value, 4)`, $W_{bat}$= `r round(bdiv$bat$statistic, 4)`). This paper therefore corroborates the existing consensus that habitat alteration can cause homogenisation without necessarily reducing richness [@Toth2014; @Dornelas2014; @McKinney2006; @Trentanovi2013; @Knop2015; @Kay2017]. 

To characterise the effect of food competition on community assembly, we considered every possible pairwise combination of species within each taxon. Each pair was classified as potentially competing (henceforth: competing) or not competing (henceforth: control) based on basic dietary guild information. Competing pairs shared dietary guilds while control pairs shared no dietary sources. A small proportion of pairs with partially overlapping diets were excluded for simplicity. Sites were classified as altered or intact based on Ecological Register metadata [@Supplement]. The intersection of these two variables produced four experimental groups: competing pairs in intact sites, control pairs in intact sites, competing pairs in altered sites, and control pairs in altered sites. We fit a hierarchical Bayesian model [@Supplement] that yields group-level estimates ($g$) for the average ($\bar{\psi_{g}}$) and standard deviation ($\sigma_{g}$) of a spatial aggregation parameter ($\psi$) that indexes pairwise co-occurrence using Fisher’s noncentral hypergeometric distribution [@Fog2008]. A negative value of $\psi$ indicates spatial segregation of the species pair, whereas a value of approximately 0 indicates independent patterns of occupancy for the pair, and a positive values indicate spatial aggregation. Models were run on all pairs (“full”), then repeated those exceeding the median incidence rate in their taxon and habitat type (“common-only”). 

Competition for food affected the spatial community structure of both bats and birds, demonstrated by shifts in the modes of the posterior distributions for $\bar{\psi_{g}}$ and $\sigma_{g}$ of competing and control groups in both taxa (Figs. 2-3). Competing pairs consistently exhibited significantly higher variance than control pairs, supporting the conclusion that spatial patterns are more variable among species that share food sources.

Overall, birds tended toward repulsive assiociations while bats tended toward attractive associations. This observation represents fundamental difference in the ecology of birds and bats. Bats tend to exhibit limiting morphological similarity [@Kingston2000], share roosts [@Swift1983], display varied foraging behaviours [@Swift1983; @Denzinger2013], and micro-partition resources, habitats and foraging times [@Aldridge1987; @Nicholls]. Meanwhile, individual neotropical bird species are well known for interspecific aggression [@Jankowski2010; @Freeman2016; @Freeman2016a], separating out along elevational gradients [@Jankowski2010], and competitive exclusion [@Terborgh1975; @Remsen1995]. Even control bird pairs are far less likely to co-occur strongly than control bat pairs, suggesting that natural beta diversity and abiotic niche specialization are much more pronounced in birds than bats.

In intact habitats, we find no compelling evidence for assemblage-wide trends of competitive exclusion in competing birds and bats. Indeed, competing pairs coexist equally to or more readily than control pairs on average, demonstrated by the posterior distribution of $\psi$ being shifted left for competing compared to control pairs in both bats (Fig. 2A-B) and birds (Fig. 3A-B). Because bats tend to aggregate and birds tend to segregate on average, this translates stronger attraction of competing bats than control bats and to weaker repulsion of competing birds than control birds, on average. Competing variances are also significantly higher than the control group suggesting that mutual food sources encourage resource sharing and partitioning in the wild even in models where $\psi$ was not significantly offset, such as models including rare species. In both taxa, this relationship is lost in altered habitats. 

For bats, the offset in $\psi$ between competing and control pairs disappears in altered habitats, and variances are higher for both competing and control groups. While resource partitioning may still be occurring in some pairs, exclusion is much more common for competing pairs under altered than intact habitats. Full models (Fig. 2B) and common-only models (Fig. 2A) agree in this respect, but full models including have an additional feature: lower $\psi$ for both competing and control groups with respect to the intact baseline. For bats, natural coexistence is therefore less common in altered habitats, particularly for rarer species, and partly independent of competitive status.

For birds, the signal of resource sharing or partitioning observed in unaltered habitats is not only lost but completely upended, with control pairs exhibiting dramatically higher $\psi$ and lower variance while competing pairs remain at unaltered levels of $\psi$ (common-only models Fig. 3A) or below them (full models Fig. 3B) in altered habitats. The higher $\psi$ of control pairs (which include all the same species as competing pairs) suggests that altered habitats supply more opportunities for co-occurrence than unaltered habitats (for instance, through resource concentration or niche construction), and competing pairs are unable to capitalise on these opportunities. This causes them to instead exhibit exclusion compared to control pairs.

We re-ran our models (both “full” and “common-only”) excluding species that were not found at least once in both altered and unaltered habitats. The purpose was to discover whether (a) compositional disparities between the two habitat types are the main drivers of the observed patterns, (b) actual changes in spatial behavior of common species is different, or (c) both. There were strong similarities between these “no-turnover” models and our original models. In bats, there was very little difference between the two model sets (Fig. 2C-D), supporting hypothesis (b). In birds, the same relationships are observed within habitats, but altered pairs exhibit universally lower ψ than they do in the model sets containing turnover (Fig. 3C-D). This suggests that the opportunities for co-occurrence are amplified primarily for species that were only sampled in altered habitats (i.e. species that thrive primarily in disturbed areas) and supports hypothesis (c). However, competing pairs still exhibit exclusion in altered habitats, and this is very pronounced in both full and common-only no-turnover models, with $\psi$ significantly lower than any other group. Altered habitats cannot consistently sustain diverse sets of competing pairs, irrespective of whether the species comprising them occur in both types of habitats or primarily in altered habitats. The no-turnover models also exhibited a stronger tendency for partitioning (competing pairs have higher $\psi$) in unaltered habitats, even for species with lower occupancies. This result signals species that are better able to partition resources are more likely to be found in both habitat types, and, conversely, those that did experience competitive exclusion in unaltered habitats are less likely to be sampled in any altered habitats.

Our results confirm that the effect of diffuse food competition on community structure can be detected using co-occurrence analysis. Unlike most analyses in the recent literature, pairs that lack an arbitrarily significant associations are not removed from the analyses, but instead used to estimate the differences between groups that interact and do not interact, thus shedding light on the effect of target interaction types on the structure of communities and controlling for differences between taxa. The inclusion of potentially many non-interacting pairs in the interacting group (i.e. pairs in the same guild that do not actually compete) reduces the power of the model significantly, but substantial differences between the groups can nonetheless be detected as we demonstrate here.

We find differences in the community assembly of bats and birds in intact habitats, but both species exhibited a preference for coexistence in competing pairs with respect to control pairs. Natural patterns changed under habitat alteration, even though richness, turnover, and composition analyses were largely unable to identify significant differences. Altered habitats afford fewer opportunities for the coexistence of bats but more opportunities for the coexistence of birds, especially those sampled exclusively in altered habitats. Competing pairs, however, exhibited lower co-occurrence than control pairs in both taxa, suggesting that altered habitats are unable to locally support rich communities of competing species the way intact habitats do. The no-turnover analyses suggests that these effects are largely attributable to changes in behaviour, and not merely to differences in species composition. The loss of competitive interactions in altered sites represents a loss of ecosystem functionality. Continuing loss of functionality could jeopardize ecosystem services that rely on competition or on the persistence of the non-dominant species in competitive interactions. Overall, habitat alteration reduces or reverses the effect of competition on community structure in both bats and birds. Our results shed light on the elusive effect of biotic interactions on community assembly and our approach helps pave the way for future research seeking to understand how interaction outcomes change across various contexts and large spatial scales.

__Acknowledgments:__ We thank the Macquarie University paleobiology lab for discussions that improved this paper. 
__Funding:__ ABT was partially supported by an IMQRES scholarship at Macquarie University.
__Author contributions:__ ABT and JA designed the study, JA curated the data, ABT analyzed data, produced the figures, and wrote the paper. ABT and APA designed, coded, and tested the model. SKL helped design supporting analyses. All authors edited the manuscript.
__Competing Interests:__ The authors declare no competing interests.
__Data and materials availability:__ All R workflows and cleaned datasets used for this analysis are available at https://github.com/anikobtoth/HabitatAlteration.

### Figures

```{r echo=FALSE, include = T, fig.height=4, fig.width=9, ord}
# Ordination
dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "sample.no"))

#PCoA plot
  p1 <- ggplot(ord[[1]], aes(x = X1, y = X2, col = status)) + geom_point() + labs(x = "Comp1", y = "Comp2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77"))
  
  p2 <- ggplot(ord[[2]], aes(x = X1, y = X2, col = status)) + geom_point() + labs(x = "Comp1", y = "Comp2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77"), labels = c("Altered", "Intact")) 
  
  p <- plot_grid(p1+theme(legend.position = "none"), 
                 p2+theme(legend.position = "none", 
                axis.title.y = element_blank()), ncol = 2, labels = c("A","B"), align = "lr")
  plot_grid(p, get_legend(p1), rel_widths = c(1, .2))

```

__Figure 1.__ Principal coordinates analysis of altered and unaltered sites for bats (A) and birds (B). There is no significant difference between the composition of altered and unaltered sites for either taxon, based on a permutation test of the site type centroids (p-values = `r w[1]` and `r w[2]` for bats and birds, respectively). 

```{r echo=FALSE, include = T}
colors <- c("#FFC2AD", "#DF8768", "#83C5BE", "#006D77")

specs <- list(geom_density(lwd = 1, alpha = .4),
              facet_wrap(parameter~., scales = "free_x"), 
              scale_color_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")), 
              scale_fill_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")))


```

```{r echo=FALSE, fig.height=4, fig.width=7, include= T}
load("../Results/Omega/interaction/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
a1 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs
load("../Results/Omega/interaction/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
a2 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
a3 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
a4 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")

legend <- get_legend(a1+ theme(legend.box.margin = margin(0, 0, 0, 7)))

p1 <- plot_grid(a1 + theme(legend.position = "none"), a2, a3, a4, align = "hv", nrow = 2, labels = LETTERS[1:4])
plot_grid(p1, legend, nrow = 1, rel_widths = c(2, 0.8))

```

__Figure 2.__ Results of bat models, depicting $\psi$ and $\sigma_{g}$ posteriors for (A) all bats with occurrences exceeding the median occurrence level in its respective habitat (B) all bats, (C) all bats with occurrences exceeding the median occurrence that occur in both intact and altered habitats, and (D) all bats that occurred in both intact and altered habitats. 

```{r echo=FALSE, fig.height=4, fig.width=7, include = T}

load("../Results/Omega/interaction/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
a1 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs 
load("../Results/Omega/interaction/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
a2 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
a3 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
a4 <- ggplot(out %>% filter(!parameter %in% c("deviance")), aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + 
  specs + theme(legend.position = "none")

legend <- get_legend(a1+ theme(legend.box.margin = margin(0, 0, 0, 7)))

p1 <- plot_grid(a1 + theme(legend.position = "none"), a2, a3, a4, align = "hv", nrow = 2, labels = LETTERS[1:4])
plot_grid(p1, legend, nrow = 1, rel_widths = c(2, 0.8))


```

__Figure 3.__ Results of bird models, depicting $\psi$ and $\sigma_{g}$ posteriors distributions. Panels as in Fig. 2.

## References



