---
title: Competitive interactions are extirpated from human-altered habitats in the Neotropics
output:
  bookdown::word_document2: default
bibliography: References.bib
csl: science.csl

---

__Authors:__ Anikó B. Tóth^1^*, Andrew P. Allen^2^, S. Kathleen Lyons^3^, John Alroy^2^

__Affiliations:__

^1^ Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW, Sydney, NSW 2052, Australia.

^2^ Department of Biological Sciences, Macquarie University, New South Wales 2109, Australia.

^3^ School of Biological Sciences, University of Nebraska-Lincoln, Lincoln, Nebraska 68588, USA.

*Correspondence to aniko.toth@unsw.edu.au

__Abstract:__ Disruption of biotic interactions by habitat destruction dismantles communities even at broad spatial scales, jeopardizing ecosystem functions. However, large scale, assemblage-wide studies of this phenomenon are uncommon and beset by logistical hurdles. Here, we use a novel approach to assess Neotropical bird and bat assemblages, asking whether spatial patterns of co-occurrence among species pairs that compete for food resources differ between altered and intact habitats. We find that competing species pairs of both taxa are more spatially segregated on average in altered habitats than intact habitats. The shift is not driven by compositional differences but by changes in spatial relationships of species. Our results suggest that altered habitats fail to support diverse competitive interactions, reversing patterns observed in the wild.

__One sentence summary:__ Coexistence of competing species of Neotropical bats and birds is reduced in human-altered habitats.

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(cowplot)
library(tidyverse)
library(igraph)
library(reshape2)
library(vegan)
library(sp)
library(stringi)
library(parallel)
library(magick)

knitr::opts_chunk$set(include = FALSE)

## Load helper functions
source('../Code/HelperFunctions.R')
## Prep raw data
source('../Code/Data_Prep.R')

```


__Main text:__
Human societies rely on ecosystem services, many of which depend on a balance of interspecific interactions that both affect and are affected by spatial patterns of species co-occurrence. For example, population control of a destructive pest species or disease vector by a predator or competitor species requires that both species occur in the same place. Human disturbance may alter or rewire interactions in unpredictable ways [@Bartley2019] and this can have detrimental consequences for ecosystems and people. However, empirical studies quantifying how biotic interactions change under degraded conditions are uncommon. Evidence that human disturbance can decouple predator-prey interactions [@Rodewald2011], increase competition for pollinators [@Holzschuh2011], and simplify or break apart interaction networks [@Haddad2000; @DeAssisBomfim2018] has been gathered mainly at local scales on systems with only a few species due to logistical hurdles. Broad patterns documenting changing interactions in entire assemblages have rarely been revealed, with the possible exception of systems where interaction data has been laboriously compiled over decades, such as plant-pollinator networks, and these are nonetheless beset by sampling inconsistencies (e.g., [@Dore2021]). 

Studying biotic interactions is particularly challenging, yet as one of the four main drivers of species distributions  [@McGill2010], it is crucial to understand how assemblage-level interaction dynamics change in various ecological contexts  [@Song2020] and in response to disturbances, to facilitate conservation of species and ecosystems in an uncertain future. For instance, species pairs can exhibit stable or unstable population dynamics depending on the presence or absence of an apex predator in the system [@Karakoc2020], or switch from facilitation to competition along a climate gradient [@Bimler2018]. Competitive interactions can result in coexistence or exclusion, depending on fitness, niche disparity, and priority effects of the species involved [@Grainger2019]. Human habitat alteration has strong potential to change these and other ecological contexts (e.g., by extirpating apex predators, or changing the climate, or concentrating resources) and shift the dynamics of biotic interactions, often without necessarily adding or removing species [@Tylianakis2017].

Although this interpretation is still contentious [@McGill2010; @Araujo2013; @Ezard2016], it is increasingly recognized that biotic interactions can scale up to influence community assembly at much larger spatiotemporal scales  than that of individual encounters [@Araujo2007; @Pollock2014; @Fraser2020]. For example, competition and facilitation can influence the modern distribution of taxa at landscape scales [@Gotelli2010], at continental scales over millennia [@Toth2019], and drive the diversification and extinction of entire clades over millions of years [@Sepkoski1984; @Alroy2010; @Fraser2020]. Human alteration of interactions can therefore have far-reaching consequences.

```{r, include = FALSE}
###### Richness - run using raw abundance data ####
# can only be run on raw data because it requires a singleton count of abundances.
# cJ1
cJ1 <- map(PAn, map_dbl, cJ1rich) %>% map(~split(., names(.) %in% unalt_sites)) %>% 
  map(map, cbind) %>% map(map, data.frame) %>% map(bind_rows, .id = "status") %>% 
  bind_rows(.id = "taxon") %>% setNames(c("taxon", "status", "richness"))
cJ1$status <- plyr::revalue(cJ1$status, c("FALSE" = "Altered", "TRUE" = "intact"))
# summary stats
cJ1 %>% group_by(taxon, status) %>% summarise(mean.rich = mean(richness), median.rich = median(richness))

# test for significant difference between altered and intact
cJ1 %>% group_by(taxon) %>% summarise(w=wilcox.test(richness~status, paired=FALSE)$p.value)

# CHAO 1
chao <- map(PAn, map_dbl, chao1) %>% map(~split(., names(.) %in% unalt_sites)) %>% 
  map(map, cbind) %>% map(map, data.frame) %>% map(bind_rows, .id = "status") %>% 
  bind_rows(.id = "taxon") %>% setNames(c("taxon", "status", "richness"))
chao$status <- plyr::revalue(chao$status, c("FALSE" = "Altered", "TRUE" = "intact"))
#summary status
chao %>% group_by(taxon, status) %>% summarise(mean.rich = mean(richness, na.rm = T), median.rich = median(richness, na.rm = T))
#significance test
chao %>% group_by(taxon) %>% summarise(w=wilcox.test(richness~status, paired=FALSE)$p.value)

```

Here we show how human habitat alteration influences one type of interspecific interaction (food competition) through its effects on spatial associations of species pairs (meaning aggregation or co-occurring more often than expected vs. segregation, co-occurring less often than expected). Our statistical modeling approach can elucidate how associations between individual pairs of species collectively result in assemblage-level spatial patterns. We apply the model to assemblages of Neotropical bats and birds downloaded from the Ecological Register [@Alroy2015] consisting of `r sum(PAn[["bat"]])` occurrences of bats and `r sum(PAn[["bird"]])` occurrences of birds at `r ncol(PAn[["bat"]])` and `r ncol(PAn[["bird"]])` sites, respectively, representing two levels of degradation (altered vs. intact habitats; Fig. S1). Our analysis demonstrates that habitat degradation influences spatial patterns of co-occurrence among competitors at regional to continental scales by altering – or even reversing – the average outcome of food competition across regional assemblages. Conventional methods of analyzing community structure are either unable to detect differences in patterns between altered and intact sites or do so without necessarily providing insight about assembly processes.

```{r matchbiogeo, echo = FALSE}
#### Match Biogeography ####
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites])) # separate intact
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites])) # and altered sites

coords <- sitedat %>% select(taxon, p.sample, latitude, longitude) %>% distinct() %>% split(.$taxon)

# Remove intact or altered sites that are not near a site of the other type. This is done to ensure the two sets have similar biogoegraphical distributions.
coords1 <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2 <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])
keep <- map2(coords1, coords2, matchbiogeo) %>% map(unlist)

PAn <- map2(PAn, keep, function(x, y) return(x[,as.character(y)]))
PAn <- map(PAn, clean.empty, minrow = 1) # remove any species that now have no occurrences

# recalculate alt/unalt split from new PAn
PAnu <- PAn %>% map(~return(.[,colnames(.) %in% unalt_sites]))
PAna <- PAn %>% map(~return(.[,!colnames(.) %in% unalt_sites]))

# coords of sites we kept, for plotting later.
coords1.keep <- purrr::map2(coords, PAnu, function(x, y) x[x$p.sample %in% colnames(y),])
coords2.keep <- purrr::map2(coords, PAna, function(x, y) x[x$p.sample %in% colnames(y),])

```


```{r betacomp, echo=FALSE, message=FALSE, warning=FALSE}

#### Beta diversity and composition analyses #####
PAnb<- tobinary(PAn)

PA <- PAnb  ## can run analyses with binary or abudance data

# with bray-curtis index
dist <- map(PA, ~t(.)) %>% map(vegdist, method = "bray")
# with Ochiai index
dist <- map(PA, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.))

# beta diversity using betadisper
beta <- map2(dist, PAn, function(x, y) betadisper(x, group = colnames(y) %in% unalt_sites) %>% anova)

# compositional change using adonis
data <- map(PA, ~t(.x) %>% data.frame() %>% rownames_to_column("p.sample") %>% mutate(ID = colnames(.x) %in% unalt_sites))
comp1 <- map2(dist, data, function(x, y) adonis(x~ID, data = y, permutations = 10000))

# compositional change using canonical correspondence analysis
dat <- map(data, select, contains("_"))
comp2 <- map2(dat, data, function(x, y) cca(x~ID, data = y) %>% anova.cca)

```

Specifically, intact and altered habitats did not differ significantly in species richness [@Supplement] or beta diversity quantified by multivariate dispersion [@Anderson2006] with two different dissimilarity metrics (Table S1). Permutational multivariate ANOVA [@Anderson2001] and canonical correspondence analysis [@Legendre2012], which are used to compare species composition, indicated a shift in bird community composition. The latter analysis indicated a shift in bat composition only when abundance rather than occurrence data were included (Table S1). To visualise differences in community composition, we ran a multidimensional scaling analysis which demonstrates that habitat alteration is unlikely to be a strong driver of compositional dissimilarity across sites (Fig. 1). Though some of these classical analyses did detect differences in community patterns between altered and intact sites, they may require abundance data and are not necessarily informative about shifts in processes.

To characterise the effect of food competition on community assembly, we considered every possible pairwise combination of species within each taxon. Each pair was classified as potentially competing (henceforth: competing) or not competing (henceforth: control) based on basic dietary guild information (Fig. 2). Competing pairs shared dietary guilds while control pairs shared no dietary sources. For simplicity, we excluded a small proportion of pairs with partially overlapping diets. Sites were classified as altered or intact based on Ecological Register metadata [@Supplement]. The intersection of these two variables produced four experimental groups: competing pairs in intact sites, control pairs in intact sites, competing pairs in altered sites, and control pairs in altered sites. Though sharing diet guilds clearly does not guarantee that a species pair competes for food, this design ensures that all pairs that do compete are classified into the same group. The inclusion of potentially many non-interacting pairs in the interacting group reduces the power of the model substantially, but differences between the groups will be conservative estimates. We fit a hierarchical Bayesian model [@Supplement] that yields group-level estimates ($g$) for the average ($\bar{\theta_{g}}$) and standard deviation ($\sigma_{g}$) of a spatial aggregation parameter ($\theta$) that indexes pairwise co-occurrence using Fisher’s noncentral hypergeometric distribution [@Fog2008]. A negative value of $\theta$ indicates spatial segregation of the species pair, whereas a value of approximately 0 indicates independent patterns of occupancy for the pair, and a positive value indicates spatial aggregation. Models were run on all pairs (“full”), then repeated those exceeding the median incidence rate in their taxon and habitat type (“common-only”) to reduce uncertainty from high numbers of pairs with few occurrences. Posterior distributions of competing pairs were compared to the control group in each habitat for each taxon using 95% HDPIs on 1,200 posterior samples (Figs 3-4, Table S2). 

Overall, birds tended to exhibit segregation while bats tended to aggregate (Figs 3-4). This observation represents fundamental difference in the ecology of birds and bats. Bats tend to exhibit limiting morphological similarity [@Kingston2000], share roosts [@Swift1983], display varied foraging behaviours [@Swift1983; @Denzinger2013], and micro-partition resources, habitats, and foraging times [@Aldridge1987; @Nicholls]. By contrast, neotropical bird species are well known for interspecific aggression [@Jankowski2010; @Freeman2016; @Freeman2016a], separating out along elevational gradients [@Jankowski2010], and competitive exclusion [@Terborgh1975; @Remsen1995]. Even control bird pairs are far less likely to aggregate strongly than control bat pairs, suggesting that natural beta diversity and abiotic niche specialization are much more pronounced in birds than bats. Nonetheless, competition for food affected the spatial community structure of both bats and birds in both altered and intact habitats as demonstrated by offsets in the posterior distributions for $\bar{\theta{g}}$ and/or $\sigma_{g}$ of competing and control groups in both taxa (Figs. 3-4, Table S1). Competing pairs consistently exhibited substantially higher variance than control pairs (no overlap of posterior intervals), supporting the conclusion that spatial patterns are more variable among species that share food sources (Fig. S2).

In intact habitats, we found no compelling evidence for assemblage-wide incidence of competitive exclusion in competing birds and bats. Indeed, competing pairs coexisted equally to or more readily than control pairs on average, demonstrated by the posterior distribution of $\bar{\theta{g}}$ being similar or shifted right for competing compared to control pairs in both bats (Fig. 3A-B) and birds (Fig. 4A-B). Because bats tend to aggregate and birds tend to segregate on average, this translated into stronger attraction of competing bats than control bats and to weaker repulsion of competing birds than control birds. Competing variances were also higher than in the control group, suggesting that mutual food sources encourage resource sharing and partitioning in the wild. This was seen even in models where $\bar{\theta{g}}$ posterior intervals overlapped, such as models including rare species. In both taxa, this relationship is lost in altered habitats.

For bats, the offset in $\bar{\theta{g}}$ between competing and control pairs disappeared in altered habitats, and variances were higher for both competing and control groups. While aggregation may still be occurring in some pairs, exclusion was much more common for competing pairs under altered than intact habitats. Full models (Fig. 3A) and common-only models (Fig. 3B) agreed in this respect, but full models have an additional feature: lower $\bar{\theta{g}}$ for both competing and control groups with respect to the intact baseline. For bats, natural coexistence was therefore less common in altered habitats, particularly for rarer species, and partly independent of competitive status.

For birds, the signal of resource sharing or partitioning observed in intact habitats is not only lost, but completely upended, with control pairs exhibiting dramatically higher $\bar{\theta{g}}$ and lower variance, while competing pairs remain at intact levels of $\bar{\theta{g}}$ (common-only models Fig. 4B) or below them (full models Fig. 4A) in altered habitats. The higher $\bar{\theta{g}}$ of control pairs suggests that altered habitats supply more opportunities for co-occurrence than intact habitats (for instance, through resource concentration or niche construction), and competing pairs are unable to capitalise on these opportunities. This causes them to instead exhibit exclusion compared to control pairs.

We re-ran our models (both “full” and “common-only”) excluding species that were not sampled in both altered and intact habitats (we will refer to these as no-turnover models). The purpose was to discover whether the main drivers of the observed patterns are (a) compositional disparities between the two habitat types, (b) actual changes in the spatial behavior of common species between the habitat types, or (c) both. There were strong similarities between these no-turnover models and our original models. In bats, there was very little difference between the two model sets (compare Fig. 3C-D to Fig. 3A-B), supporting hypothesis (b). In birds, the same relationships are observed within habitats, but pairs in intact habitats generally exhibit higher $\bar{\theta{g}}$ (Fig. 4C-D) compared to the original model sets (Fig. 4A-B), both in absolute terms and with respect to altered pairs. This suggests that the opportunities for aggregation are amplified especially for species that were only sampled in altered habitats (i.e., species that thrive primarily in disturbed areas) and supports hypothesis (c). However, competing pairs still exhibit exclusion in altered habitats, and this is very pronounced in both no-turnover models, with $\bar{\theta{g}}$ of competing pairs in altered habitats significantly lower than in any other group. Altered habitats cannot consistently sustain diverse sets of competing pairs, irrespective of whether the species comprising them occur in both types of habitats or primarily in altered habitats. The no-turnover models also exhibit a stronger tendency for partitioning (competing pairs have higher $\bar{\theta{g}}$) in intact habitats, even for species with lower occupancies. This result suggests species that are better able to partition resources are more likely to be found in both habitat types, and, conversely, those that did experience competitive exclusion in intact habitats are less likely to be sampled in any altered habitats.

We find differences in the community structures of bats and birds in intact habitats, but both taxa exhibited higher coexistence in competing pairs with respect to control pairs, suggesting that resource partitioning occurs most frequently instead of competitive exclusion. Natural patterns changed in altered habitats, even though richness, turnover, and composition were largely similar. Altered habitats afford fewer opportunities for the coexistence of bats but more opportunities for the coexistence of birds, especially those sampled exclusively in altered habitats. Competing pairs, however, exhibited equal (bats) or lower (birds) coexistence than control pairs under altered conditions. Thus, altered habitats are unable to consistently support rich local communities of competing species through partitioning as seen in intact habitats. The no-turnover analyses indicate that these effects are largely attributable to changes in spatial behaviour, and not merely to differences in species composition.

Overall, habitat alteration reduces or reverses the effect of competition on community structure in both bats and birds. This loss of competitive interactions in altered sites may represent a loss of ecosystem functionality. Continuing loss of functionality could jeopardize ecosystem services that rely on competition or on the persistence of the less dominant species in competitive interactions. Our results add to mounting evidence that human activity changes drivers of community structure without directly adding or removing species, and further suggest that this can play out over landscape scales. Our results shed light on the elusive effect of biotic interactions on community assembly, and our approach helps pave the way for future research seeking to understand how interaction outcomes change across various contexts and at large spatial scales.


__Acknowledgments:__ We thank the Macquarie University paleobiology lab for discussions that improved this paper. 

__Funding:__ ABT was partially supported by an IMQRES scholarship at Macquarie University.

__Author contributions:__ ABT and JA designed the study, JA curated the data, ABT analyzed data, produced the figures, and wrote the paper. ABT and APA designed, coded, and tested the model. SKL helped design supporting analyses. All authors edited the manuscript.

__Competing Interests:__ The authors declare no competing interests.

__Data and materials availability:__ All R workflows and cleaned datasets used for this analysis are available at https://github.com/anikobtoth/HabitatAlteration.

### Figures

```{r echo=FALSE, fig.height=4, fig.width=9, message=FALSE, warning=FALSE, include=T}

# Ordination
dist <- map(PAn, ochiaiMatrix) %>% map(as.dist, upper = F) %>% map(~return(1-.)) 
#dist <- map(PA, ~t(.)) %>% map(vegdist, method = "bray")
ord <- map(dist, cmdscale, k = 2) %>% map(data.frame)
ord <- map(ord, ~merge(., y = sitedat, all.x = T, all.y = F, by.x = 0, by.y = "p.sample"))

bat <- image_read("../../documents/bat_pic.png") %>% image_fill('none') %>% as.raster()
bird <- image_read("../../documents/bird_pic.png") %>% image_fill('none') %>% as.raster()

#PCoA plot
  p1 <- ggplot(ord[[1]], aes(x = X1, y = X2, col = status)) + geom_point(size = 2.6) + labs(x = "PCoA 1", y = "PCoA 2", col = "alteration") + theme(axis.text = element_text(size = 12), legend.text = element_text(size = 12), legend.title =  element_text(size = 12, face = "bold")) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77")) +
    annotation_raster(bat, -0.45, -0.2, 0.27, 0.42)#+ geom_point(aes(x = mean(X1), y = mean(X2), col = status), pch = 3, size = 3)
  
  p2 <- ggplot(ord[[2]], aes(x = X1, y = X2, col = status)) + geom_point(size = 2.6) + labs(x = "PCoA 1", y = "PCoA 2", col = "alteration") + theme(axis.text = element_text(size = 12)) + labs(col = "status") + scale_colour_manual(values = c("#DF8768", "#006D77"), labels = c("Altered", "Intact")) +
     annotation_raster(bird, -0.45, -0.25, 0.25, 0.44) #+ geom_point(aes(x = mean(X1), y = mean(X2), col = status), pch = 3, size = 3)
  
  p <- plot_grid(p1+theme(legend.position = "none"), 
                 p2+theme(legend.position = "none", 
                axis.title.y = element_blank()), ncol = 2, labels = c("A","B"), align = "lr")
  plot_grid(p, get_legend(p1), rel_widths = c(1, .2))

```

__Figure 1.__ Principal coordinates analysis of altered and intact sites for bats (A) and birds (B). There is no significant difference between the composition of altered and intact sites for either taxon, based on a permutation test of the site type centroids (p-values = `r w[1]` and `r w[2]` for bats and birds, respectively). 


__Figure 2.__ Conceptual figure demonstrating the experimental design. (A) All pairwise combinations of species are classified as potentially competing or not competing for food based on basic dietary information. (B) A distribution of $\psi$ is calculated for each pair, shown as an adjacency matrix. From these distributions, (C) the cumulative posterior distribution of group means $\bar{\theta{g}}$ is derived and compared. 

```{r echo=FALSE, include = T, warnings = FALSE}
colors <- c("#FFC2AD", "#DF8768", "#83C5BE", "#006D77")
parameter.labels <- c(omega = "group θ", sigma = "variance (σ)")
ann <- data.frame(value = -Inf, lab = LETTERS[1:4], parameter = "omega", model = unique(out$model), status_dietmatch = "intact_Different")

specs <- list(geom_density(lwd = 1, alpha = .4),
              facet_grid(model~parameter, scales = "free_x", labeller = labeller(parameter = parameter.labels)), 
              scale_color_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")), 
              scale_fill_manual(values = colors, labels = c("Altered control", "Altered competing", "Intact control", "Intact competing")), 
              labs(fill = "legend", col = "legend"), 
              theme(text = element_text(size = 12)),
              geom_text(data = ann, y = Inf, label = ann$lab, col = "black", fontface = 2, hjust = -1, vjust = 1)
              )



```

```{r echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE, include=T}
load("../Results/Omega/interaction/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
bat_common <- out
load("../Results/Omega/interaction/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
bat_full <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianTRUE_100reps_interaction_baggingFALSE.RData")
bat_common_noTurnover <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bat_medianFALSE_100reps_interaction_baggingFALSE.RData")
bat_full_noTurnover <- out

out <- bind_rows(list(full = bat_full, 
                      common = bat_common, 
                      full_noTurnover = bat_full_noTurnover, 
                      common_noTurnover = bat_common_noTurnover), .id = "model") %>% filter(!parameter == "deviance")

out$model <- factor(out$model, levels = c("full", "common", "full_noTurnover", "common_noTurnover"))

ggplot(out, aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + specs


```

__Figure 3.__ $\bar{\theta{g}}$ and $\sigma_{g}$ posteriors for bat models, specifically (A) all bats, (B) all bats having occurrences that exceed the median occurrence count in the respective habitat, (C) and all bats that occurred in both intact and altered habitats, and (D) all bats having occurrences that exceed the median occurrence in both intact and altered habitats. C and D represent the models in which altered and intact faunas are composed of the same suite of species (no-turnover models).

```{r echo=FALSE, fig.height=7, fig.width=7, include = T, warning = FALSE, message = FALSE}

load("../Results/Omega/interaction/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
bird_common <- out
load("../Results/Omega/interaction/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
bird_full <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianTRUE_100reps_interaction_baggingFALSE.RData")
bird_common_noTurnover <- out
load("../Results/Omega/interaction/shared/out_jagsfit_bird_medianFALSE_100reps_interaction_baggingFALSE.RData")
bird_full_noTurnover <- out


out <- bind_rows(list(full = bird_full, 
                      common = bird_common, 
                      full_noTurnover = bird_full_noTurnover, 
                      common_noTurnover = bird_common_noTurnover), .id = "model") %>% filter(!parameter == "deviance")
out$model <- factor(out$model, levels = c("full", "common", "full_noTurnover", "common_noTurnover"))

ggplot(out, aes(x = value, col = status_dietmatch, fill = status_dietmatch)) + specs


```

__Figure 4.__ $\bar{\theta{g}}$ and $\sigma_{g}$ posteriors distributions for bird models. Panels as in Fig. 2.

## References



